---
title:  "Divergence across N-MT genes"
author: "Natalia Zajac"
output:
  html_document: 
  highlight: pygments
theme: sand
code_folding: hide
toc: yes
toc_depth: 4
editor_options: 
  chunk_output_type: inline
---

Started on `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`

```{r setup, include=FALSE}
library(poolfstat)
library(pheatmap)
library(dplyr)
library(cowplot)
library(patchwork)
library(readr)
library(stringr)
library(tidyverse)
library(textshaping)
library(ggVennDiagram)
library(ComplexHeatmap)
library(data.table)
library(ezRun)
library(scales)
library(GSEABase)
library(tidyverse)
library(rstatix)
library(ggpubr)
library(ggrepel)
library(GO.db)
library(enrichR)
library(clusterProfiler)
working_dir=getwd()

##Load the dataframe of Fst per window (together with nucleotide diversity per window, results from Fisher's exact test and results from BayPass)
Fst_grendalf = read_delim(paste0(working_dir, "../FST/Popoolation2/Sliding_Fst_per_Lake/Fst_sliding.Grendalf.Pi.FET.C2.txt"))

df = read_delim(paste0(working_dir,"/../test.sync")) # just for colnames
poolnames = str_remove(sapply(str_split(colnames(df)[c(-1:-3)], "\\."), .subset, 2), "B-")
metadata = data.frame(pool = poolnames , lake = sapply(str_split(poolnames, "_"), .subset, 1)) %>% column_to_rownames("pool") %>% mutate(lake = if_else(lake == "Alex", "Alexandrina", lake))

colors = list(lake = setNames(c("orange", "cornflowerblue","gold", "purple", "darkgreen"), unique(metadata$lake)))

## Load gene annotation and merge with the previous dataframe - annotation created with Genome_Assembly_Organization/GCA_013407085.1.R
geneAnno = read_delim(paste0(working_dir,"/../Atriophallophorus_winterbourni/EBI/GCA_013407085.1/Annotation/Genes/genes_annotation_byGene.txt"))

genes = data.table(geneAnno[,c(6,8,9)])
colnames(genes) = c("Chr", "start", "end")
setkey(genes, Chr, start, end)

snps = Fst_grendalf[,c(1,2,3)] %>% unique() 
snps = data.table(snps)
setkey(snps, chrom, start, end)
obj = foverlaps(snps, genes, by.x = c("chrom", "start", "end"), 
                by.y = c("Chr", "start", "end"), type = "within", nomatch = 0)
obj = merge(data.frame(obj), geneAnno[,c(3,6,8,9)], by.x = c("chrom", "start", "end"), by.y = c("seqid", "start", "end"), all = TRUE)
Fst_grendalf = merge(Fst_grendalf, obj %>% select(-2,-3) , by.x = c("chrom", "start", "end"), by.y = c("chrom", "i.start", "i.end"), all.x = TRUE) %>% replace(is.na(.), "")
Fst_grendalf$window = paste0(Fst_grendalf$chrom, "-", Fst_grendalf$start, "-", Fst_grendalf$end)

##Create Gene Ontology Annotation df list
term2gene = NULL
term2gene[['BP']] = geneAnno[,c(3,12)] %>% filter(!is.na(`GO BP`)) %>%
    separate_rows(`GO BP`, sep = "; ") %>%
    dplyr::rename(Term = `GO BP`, Gene = gene_name) %>% dplyr::select(c(Term,Gene))
term2gene[['MF']] = geneAnno[,c(3,13)] %>% filter(!is.na(`GO MF`)) %>%
    separate_rows(`GO MF`, sep = "; ") %>%
    dplyr::rename(Term = `GO MF`, Gene = gene_name) %>% dplyr::select(c(Term,Gene))
term2gene[['CC']] = geneAnno[,c(3,14)] %>% filter(!is.na(`GO CC`)) %>%
    separate_rows(`GO CC`, sep = "; ") %>%
    dplyr::rename(Term = `GO CC`, Gene = gene_name) %>% dplyr::select(c(Term,Gene))

term2name= NULL
term2name[["BP"]] = data.frame(Term = term2gene$BP$Term, Name = sapply(term2gene$BP$Term, function(x){Term(x)[[1]]}))
term2name[["MF"]] = data.frame(Term = term2gene$MF$Term, Name = sapply(term2gene$MF$Term, function(x){Term(x)[[1]]}))
term2name[["CC"]] = data.frame(Term = term2gene$CC$Term, Name = sapply(term2gene$CC$Term, function(x){Term(x)[[1]]}))

## Load information content per GO
IC = read.delim(paste0(working_dir,"/../Atriophallophorus_winterbourni/EBI/GCA_013407085.1/GO_files/Atrio_GO_IC"), header = F, sep = " ")

#Add KO annotation
kegg = read.delim("../Gene2Kegg_Awinterbourni.txt",header = F, sep = "\t")
kegg = kegg[-1,c(1,2,3)]
term2gene$Kegg = kegg[kegg$V2 != "",c(2,1)]
colnames(term2gene$Kegg) = c("KO","Gene")
term2gene$Kegg$Gene = str_remove(term2gene$Kegg$Gene, "-mRNA-1")
term2name$Kegg = kegg[kegg$V2 != "",c(2,3)]
colnames(term2name$Kegg) = c("KO", "Name")

## Compute average per window of each two-lake comparison
for_plots = NULL
for (i in unique(Fst_grendalf$label)[c(2:5,7:10,11,15)]){
  tb = Fst_grendalf[Fst_grendalf$label == i,]
  x = list("Grendalf" = tb[tb$MeanGrendalfFst_4sd == "outlier",]$window, 
           "C2" = tb[tb$C2std_Sig_SNPs > 0,]$window, 
           "FetMean" = tb[tb$FET_SigMean_SNPs > 0,]$window, 
           "FetMedian" = tb[tb$FET_SigMedian_SNPs > 0,]$window)
  tb$color = if_else(tb$window %in% Reduce(intersect, x), "final_outlier", "not_outlier")
  for_plots = rbind(for_plots, tb)
}

for_plots = for_plots %>% mutate(label2 = case_when(label %in% c("Alexandrina_Paringa", "Alexandrina_Mapourika", "Middleton_Paringa", "Mapourika_Middleton") ~ "West vs South Canterbury", 
                                        label %in% c("Alexandrina_Selfe", "Middleton_Selfe") ~ "South Canterbury vs Selfe", 
                                        label %in% c("Mapourika_Selfe", "Paringa_Selfe") ~ "West vs Selfe", 
                                        label == "Alexandrina_Middleton" ~ "East vs East",  label == "Mapourika_Paringa" ~ "West vs West"))

#Annotate Gos with GOSSlims
og_slimdf = NULL
for (i in c("BP", "MF", "CC")){
  myIds <- term2gene[[i]]$Term
  myCollection <- GOCollection(myIds)
  fl <- paste0(working_dir,"/../Atriophallophorus_winterbourni/EBI/GCA_013407085.1/GO_files/goslim.obo") #go_slim_generic
  slim <- getOBOCollection(fl)
  slimdf <- goSlim(myCollection, slim, ontology = i)
  if (i == "BP"){
    GO.db::GOBPOFFSPRING
      mappedIds <-
        function(df, collection, OFFSPRING)
        {
        map <- as.list(OFFSPRING[rownames(df)])
        mapped <- lapply(map, intersect, ids(collection))
        df[["go_terms"]] <- vapply(unname(mapped), paste, collapse = ";", character(1L))
        df
        }
      slimdf <- mappedIds(slimdf, myCollection, GOBPOFFSPRING)
  } else if (i == "MF"){
     GO.db::GOMFOFFSPRING
      mappedIds <-
        function(df, collection, OFFSPRING)
          {
          map <- as.list(OFFSPRING[rownames(df)])
          mapped <- lapply(map, intersect, ids(collection))
          df[["go_terms"]] <- vapply(unname(mapped), paste, collapse = ";", character(1L))
          df
          }
      slimdf <- mappedIds(slimdf, myCollection, GOMFOFFSPRING)
  } else {
    GO.db::GOCCOFFSPRING
      mappedIds <-
        function(df, collection, OFFSPRING)
          {
          map <- as.list(OFFSPRING[rownames(df)])
          mapped <- lapply(map, intersect, ids(collection))
          df[["go_terms"]] <- vapply(unname(mapped), paste, collapse = ";", character(1L))
          df
          }
      slimdf <- mappedIds(slimdf, myCollection, GOCCOFFSPRING)
  }
  slimdf = slimdf %>% rownames_to_column("GOslim") %>% separate_longer_delim(go_terms, delim = ";")
  slimdf = merge(merge(slimdf, term2gene[[i]], by.x = "go_terms", by.y = "Term"), term2name[[i]], by.x = "go_terms", by.y = "Term")
  slimdf$GO = i
  og_slimdf = rbind(og_slimdf,slimdf)
}

#OR load a precomputed one because this takes a while:
og_slimdf = read.delim("ogslimdf.txt")

##Define a useful colour scheme
Cb64k <- c("#000000", "#FFFF00", "#1CE6FF", "#FF34FF", "#FF4A46", "#008941", "#006FA6", "#A30059",
           "#FFDBE5", "#7A4900", "#0000A6", "#63FFAC", "#B79762", "#004D43", "#8FB0FF", "#997D87",
           "#5A0007", "#809693", "#FEFFE6", "#1B4400", "#4FC601", "#3B5DFF", "#4A3B53", "#FF2F80",
           "#61615A", "#BA0900", "#6B7900", "#00C2A0", "#FFAA92", "#FF90C9", "#B903AA", "#D16100",
           "#DDEFFF", "#000035", "#7B4F4B", "#A1C299", "#300018", "#0AA6D8", "#013349", "#00846F",
           "#372101", "#FFB500", "#C2FFED", "#A079BF", "#CC0744", "#C0B9B2", "#C2FF99", "#001E09",
           "#00489C", "#6F0062", "#0CBD66", "#EEC3FF", "#456D75", "#B77B68", "#7A87A1", "#788D66",
           "#885578", "#FAD09F", "#FF8A9A", "#D157A0", "#BEC459", "#456648", "#0086ED", "#886F4C",
           "#34362D", "#B4A8BD", "#00A6AA", "#452C2C", "#636375", "#A3C8C9", "#FF913F", "#938A81",
           "#575329", "#00FECF", "#B05B6F", "#8CD0FF", "#3B9700", "#04F757", "#C8A1A1", "#1E6E00",
           "#7900D7", "#A77500", "#6367A9", "#A05837", "#6B002C", "#772600", "#D790FF", "#9B9700",
           "#549E79", "#FFF69F", "#201625", "#72418F", "#BC23FF", "#99ADC0", "#3A2465", "#922329",
           "#5B4534", "#FDE8DC", "#404E55", "#0089A3", "#CB7E98", "#A4E804", "#324E72", "#6A3A4C")
```


# {.tabset}

## Selection of N-MT genes

```{r}
#https://www.mdpi.com/2218-273X/12/3/427#B164-biomolecules-12-00427

kegg_mt = read.csv(paste0(working_dir,"/../Mt_N_Confirmation/ko03029_annotated.csv"))
kegg_mtKOs = kegg_mt$KO_ID
kegg_df =  merge(term2name$Kegg[term2name$Kegg$KO %in% kegg_mtKOs,],term2gene$Kegg[term2gene$Kegg$KO %in% term2name$Kegg[term2name$Kegg$KO %in% kegg_mtKOs,]$KO,], by = "KO")
kegg_df = merge(kegg_df, kegg_mt, by.x = "KO", by.y = "KO_ID", all.x = TRUE)
kegg_df = kegg_df[,c(4,1,5,6,3)]
colnames(kegg_df) = c("Broad_ID", "Narrow_ID", "Broad_Term", "Narrow_Term", "Gene")
kegg_df$Source = "KEGG"

mtcandidates=NULL
mtcandidates[["GO_mitochondrion_organization"]] = og_slimdf[og_slimdf$Term == "mitochondrion organization",] %>% unique()
mtcandidates[["GO_mitochondrial_gene_expression"]] = og_slimdf[og_slimdf$Term == "mitochondrial gene expression",] %>% unique()
mtcandidates[["GO_mitochondrial_membrane_building_blocks"]] = og_slimdf[og_slimdf$Term == "mitochondrion",] %>% unique()
mtcandidates[["GO_mitochondrial_organelles_and_ER_interaction"]] = og_slimdf[og_slimdf$GOslim == "GO:0043226" & grepl("mitochon", og_slimdf$Name),] %>% unique()
mtcandidates[["GO_DNA_replication"]] = og_slimdf[og_slimdf$Term == "DNA replication" & og_slimdf$Name == "mitochondrial DNA replication",] %>% unique()
mtcandidates[["GO_programmed_cell_death"]] = og_slimdf[og_slimdf$Term == "programmed cell death" & og_slimdf$Name %in%  c("apoptotic mitochondrial changes", "mitochondrial outer membrane permeabilization"),] %>% unique()
mtcandidates[["GO_generation_of_precursor_metabolites_and_energy"]] = og_slimdf[og_slimdf$Name %in% c("tricarboxylic acid cycle", "oxidative phosphorylation", "electron transport chain", "mitochondrial electron transport, ubiquinol to cytochrome c", "mitochondrial electron transport, cytochrome c to oxygen", "aerobic respiration", "ATP synthesis coupled electron transport", "mitochondrial ATP synthesis coupled electron transport", "cellular respiration"),] %>% unique()
mtcandidates[["GO_lipid_metabolic_process"]] = og_slimdf[og_slimdf$Name == "fatty acid beta-oxidation",] %>% unique()
mtcandidates[["GO_proton_motive_force-driven_ATP_synthesis"]] = og_slimdf[og_slimdf$Term == "carbohydrate derivative metabolic process" & og_slimdf$Name == "proton motive force-driven ATP synthesis",] %>% unique()
mtcandidates[["GO_mitochondrial_respiratory_chain_complex_assembly"]] = og_slimdf[og_slimdf$Term == "protein-containing complex assembly" & og_slimdf$Name %in%  c("mitochondrial respiratory chain complex I assembly","mitochondrial respiratory chain complex assembly","mitochondrial respiratory chain complex III assembly"),] %>% unique()
mtcandidates[["GO_NADH_dehydrogenase_activity"]] = og_slimdf[og_slimdf$Term == "oxidoreductase activity" & og_slimdf$Name %in%  c("NADH dehydrogenase activity","NADH dehydrogenase (ubiquinone) activity","NADH dehydrogenase (quinone) activity" ),] %>% unique()

mtcandidates = lapply(mtcandidates, function(x){
  x = x[,c(1,2,5,7,6,8)]
  colnames(x) =  c("Narrow_ID", "Broad_ID", "Broad_Term", "Narrow_Term", "Gene", "Source")
  return(x)})

kegg_to_goslim = read.csv("../Mt_N_Confirmation/kegg_to_goslim_mapping.csv")

for (i in unique(kegg_to_goslim$KEGG_Level_3)){
  go_term = kegg_to_goslim[kegg_to_goslim$KEGG_Level_3 == i,]$GO_slim_match_1
  subset = kegg_df[kegg_df$Narrow_Term %in% i,]
  mtcandidates[[go_term]] = rbind(mtcandidates[[go_term]],subset)
}

mtcandidatesGenes = lapply(mtcandidates, function(x){unique(x$Gene)})
print(length(unique(unlist(unname(mtcandidatesGenes)))))
mtcandidatestoPlot = bind_rows(mtcandidates, .id = "Category")
mtcandidatestoPlot = mtcandidatestoPlot %>% group_by(Category) %>% summarise(n_distinct(Gene))

oxphos = unique(c(bind_rows(mtcandidates) %>% filter(grepl("ATP synthesis", Narrow_Term)) %>% pull(Gene) %>% unique(), bind_rows(mtcandidates) %>% filter(grepl("oxidative phosphorylation", Narrow_Term)) %>% pull(Gene) %>% unique(), bind_rows(mtcandidates) %>% filter(grepl("ATP synthase", Narrow_Term)) %>% pull(Gene) %>% unique(), bind_rows(mtcandidates) %>% filter(grepl("NADH dehydrogenase", Narrow_Term)) %>% pull(Gene) %>% unique(), bind_rows(mtcandidates) %>% filter(grepl("cytochrome c oxidase", Narrow_Term)) %>% pull(Gene) %>% unique(),bind_rows(mtcandidates) %>% filter(grepl("succinate", Narrow_Term)) %>% pull(Gene) %>% unique()))
```

Confirm how many genes are schistosoma orthologs

```{r}
# Select orthologs
orth = read_delim(paste0(working_dir, "../Mt_N_Confirmation/atriophallophorus_winterbourni.PRJNA636673.WBPS19.orthologs.tsv"))
orth = orth[orth$gene_id %in% unique(unname(unlist(mtcandidatesGenes))) & grepl("Schistosoma mansoni", orth$ortholog_species_name),]

# Load the ortholog annotations
eggnog = read.delim(paste0(working_dir,"/../Mt_N_Confirmation/mitochondrial_genes_schistosoma_orthologs.annotations_Eggnog.txt"), header = TRUE, sep = "\t")
oma = read.delim(paste0(working_dir,"/../Mt_N_Confirmation/mitochondrial_genes_schistosoma_orthologs.annotations_OMA.txt"), header = TRUE, sep = "\t")
colnames(oma) = c("ortholog_gene_id","gene_id","ortholog_species_name", "GOs")
pannzer2 = read.delim(paste0(working_dir,"/../Mt_N_Confirmation/mitochondrial_genes_schistosoma_orthologs.annotations_Pannzer2.txt"), header = TRUE, sep = "\t")
colnames(pannzer2) = c("ortholog_gene_id","gene_id","ortholog_species_name", "GOs")
all_schisto = rbind(pannzer2,oma,eggnog)

all_schisto_slimdf = NULL
for (i in c("BP", "MF", "CC")){
  myIds <- all_schisto$GOs[!is.na(all_schisto$GOs)]
  myIds = unique(myIds)
  myCollection <- GOCollection(myIds)
  fl <- paste0(working_dir,"/../Atriophallophorus_winterbourni/EBI/GCA_013407085.1/GO_files/goslim.obo") #go_slim_generic
  slim <- getOBOCollection(fl)
  slimdf <- goSlim(myCollection, slim, ontology = i)
  if (i == "BP"){
    GO.db::GOBPOFFSPRING
      mappedIds <-
        function(df, collection, OFFSPRING)
        {
        map <- as.list(OFFSPRING[rownames(df)])
        mapped <- lapply(map, intersect, ids(collection))
        df[["go_terms"]] <- vapply(unname(mapped), paste, collapse = ";", character(1L))
        df
        }
      slimdf <- mappedIds(slimdf, myCollection, GOBPOFFSPRING)
  } else if (i == "MF"){
     GO.db::GOMFOFFSPRING
      mappedIds <-
        function(df, collection, OFFSPRING)
          {
          map <- as.list(OFFSPRING[rownames(df)])
          mapped <- lapply(map, intersect, ids(collection))
          df[["go_terms"]] <- vapply(unname(mapped), paste, collapse = ";", character(1L))
          df
          }
      slimdf <- mappedIds(slimdf, myCollection, GOMFOFFSPRING)
  } else {
    GO.db::GOCCOFFSPRING
      mappedIds <-
        function(df, collection, OFFSPRING)
          {
          map <- as.list(OFFSPRING[rownames(df)])
          mapped <- lapply(map, intersect, ids(collection))
          df[["go_terms"]] <- vapply(unname(mapped), paste, collapse = ";", character(1L))
          df
          }
      slimdf <- mappedIds(slimdf, myCollection, GOCCOFFSPRING)
  }
  slimdf = slimdf %>% rownames_to_column("GOslim") %>% separate_longer_delim(go_terms, delim = ";")
  slimdf = merge(slimdf, all_schisto, by.x = "go_terms", by.y = "GOs")
  slimdf$GOdb = i
  all_schisto_slimdf = rbind(all_schisto_slimdf,slimdf)
}

length(unique(all_schisto$gene_id))/length(unique(bind_rows(mtcandidates)$Gene))

all_schisto_slimdf1 = merge(bind_rows(mtcandidates)[,c(2,5)], all_schisto_slimdf[,c(2,6,7)], by.x = c("Broad_ID", "Gene"), by.y = c("GOslim", "gene_id"))
length(unique(all_schisto_slimdf1$Gene))/length(unique(all_schisto$gene_id))

all_schisto_slimdf2 = merge(bind_rows(mtcandidates)[,c(1,2,5)], all_schisto_slimdf[,c(1,2,6,7)], by.x = c("Narrow_ID", "Broad_ID","Gene"), by.y = c("go_terms","GOslim", "gene_id"))
length(unique(all_schisto_slimdf2$Gene))/length(unique(all_schisto$gene_id))

```

## Figure 4D

```{r, fig.height=5, fig.width=13}
pdf(paste0(working_dir,"/../Figures/GeneCount_MitochondrialCategories.pdf"), height = 5, width = 13)
mtsummary = bind_rows(mtcandidates, .id = "Category") %>% group_by(Category) %>% summarise("Gene Number" = n_distinct(Gene), "GO and KEGG IDs" = n_distinct(Narrow_ID)) %>% mutate(Category = str_remove(Category,"GO_")) %>% reshape2::melt(id.vars = "Category")
function_levels = mtsummary[,c(1,3)] %>% unique() %>% arrange(Category) %>% dplyr::select(Category)  %>% unique()
function_levels = function_levels$Category
mtsummary$Category = factor(mtsummary$Category, levels = function_levels)
funcols = setNames(c(brewPalette(n=10),"brown"),levels(mtsummary$Category))

ggplot(mtsummary, aes(factor(Category, levels = function_levels), value, fill = Category)) + geom_bar(stat = "identity", position = "dodge") + theme_bw() +
    theme(axis.text.x = element_blank(), axis.title.x = element_blank()) + scale_fill_manual(values = funcols) + facet_grid(~variable, scales = "free", space = "free") + guides(fill=guide_legend(ncol=1)) + labs(fill = "N-MT category", y = "Count") + theme(text = element_text(size = 15), legend.text = element_text(size = 11), strip.text = element_text(size = 18),axis.text.y = element_text(size = 18), axis.title.y = element_text(size = 18)) + 
    geom_text(data = mtsummary, aes(x = factor(Category, levels = function_levels), y = Inf, label = value), inherit.aes = FALSE, vjust = 1.5) + scale_y_sqrt()

dev.off()
```


```{r}

xx = merge(for_plots[,c(1:5,10,11,12,13,14,18,19,20,21)], bind_rows(mtcandidates, .id = "Category")[,c(1,4,5,6,7)], by.x = "gene_name", by.y = "Gene", all.x = TRUE) %>% mutate(Category = if_else(is.na(Category) == TRUE, "Rest", Category))
xx = xx %>% mutate(Function2 = if_else(Category == "Rest", "Rest", "Mitochondrial"))
xx = xx %>% mutate(label2 = case_when(label %in% c("Alexandrina_Paringa", "Alexandrina_Mapourika", "Middleton_Paringa", "Mapourika_Middleton") ~ "West vs South Canterbury", label %in% c("Alexandrina_Selfe", "Middleton_Selfe") ~ "South Canterbury vs Selfe",  label %in% c("Mapourika_Selfe", "Paringa_Selfe") ~ "West vs Selfe",  label == "Alexandrina_Middleton" ~ "within South Canterbury",  label == "Mapourika_Paringa" ~ "within West"))
xx$label2 = factor(xx$label2, levels = c("within West", "West vs Selfe", "West vs South Canterbury", "South Canterbury vs Selfe", "within South Canterbury"))

distance = paste0(working_dir,"/../tLCP_1400.txt")
distance = read.delim(distance, sep = " ")
distance = distance[rownames(distance) %in% unique(metadata$lake),colnames(distance) %in% unique(metadata$lake)] %>% rownames_to_column() %>% reshape2::melt()
distance = distance[!duplicated(distance),]
distance$label = paste0(distance$rowname, "_", distance$variable)
distance$distance = distance$value
distance = distance[,c(5,4)]
xx = merge(xx, distance, by = "label", all.x = TRUE)
xxx = xx %>% arrange(chrom,start, end) 
```

## Figure 4B

```{r, fig.height=9, fig.width=5}

pdf(paste0(working_dir,"/../Figures/ProportionOutliers_FST_Mitochondrial.pdf"), width = 6, height = 9)
p1 = xx %>% mutate(Function2 = if_else(Category == "Rest", "Non-Mitochondrial", "Mitochondrial")) %>% group_by(color, label, label2, gene_name, Function2) %>% summarise(nb_windows = n_distinct(window)) %>% pivot_wider(names_from = color, values_from = nb_windows) %>% replace(is.na(.), 0) %>% mutate(prop_final_out = final_outlier/(final_outlier+not_outlier)) %>% group_by(label2, Function2) %>% summarise(prop_final_out = mean(prop_final_out)) %>% ggplot(aes(prop_final_out,str_wrap(label2,10), color = Function2)) + geom_point(size = 3) + labs(x = "Average proportion of outliers", y = "", color = "category") + theme(plot.margin = unit(c(5,5,5,5),"mm"), axis.text.y = element_text(size = 13), legend.text = element_text(size = 15), legend.position = "top", axis.title.x = element_text(size = 13), axis.text.x = element_text(size = 13)) + scale_color_manual(values = c("black","darkgray"))

p2 = xx %>% mutate(Function2 = if_else(Category == "Rest", "Non-Mitochondrial", "Mitochondrial")) %>% group_by(label2, Function2) %>% summarise(Fst = mean(Fst)) %>% ggplot(aes(Fst, str_wrap(label2,10),color = Function2)) + geom_point(size = 3) + labs(y = "", color = "category") + xlab(expression(F[ST])) + theme(plot.margin = unit(c(5,5,5,5),"mm"), axis.text.y = element_text(size = 13), legend.position = "", legend.text = element_text(size = 15),  axis.title.x = element_text(size = 13), axis.text.x = element_text(size = 13)) + scale_color_manual(values = c("black","darkgray"))
p1/p2
dev.off()
```

## Figure 4F

```{r, fig.height=10, fig.width=15}
mtclines = xx %>% mutate(Category = if_else(Category=="Rest", "Non-MT", Category)) %>% group_by(label2, Category) %>% summarise(Fst = mean(Fst), distance = mean(distance)) %>% filter(!label2 %in% c("West vs West", "East vs East"))
mtclines$Category = str_remove(mtclines$Category, "GO_")
mtclines$Category = factor(mtclines$Category, levels = c(unique(mtclines$Category)[!grepl("Non-MT", unique(mtclines$Category))], "Non-MT"))
funcols = c(funcols,setNames("darkgrey","Non-MT"))


line_labels <- mtclines %>% data.frame() %>%
    select(distance, label2) %>%
    distinct()

line_segments <- mtclines %>%
    arrange(Category, distance) %>%
    group_by(Category) %>%  
    mutate(
        next_distance = lead(distance),
        next_Fst = lead(Fst)
    ) %>%
    filter(!is.na(next_distance), (next_distance - distance) <= 50) %>%
    ungroup()

pdf(paste0(working_dir,"/../Figures/mtcategories_clinesv3.pdf"), width = 8, height = 7)
p1 = ggplot(mtclines, aes(x = Fst, y = distance)) +
    # Draw only short lines between points
    geom_segment(data = line_segments,
                 aes(x = Fst, xend = next_Fst,
                     y = distance, yend = next_distance,
                     color = Category),
                 linewidth = 0.8,
                 inherit.aes = FALSE) +
    geom_point(aes(color = Category), size = 2) +
    geom_hline(yintercept = unique(mtclines$distance), 
               linetype = "dashed", color = "black", alpha = 0.5) +
    geom_text(data = line_labels,
              aes(x = 0, y = distance, label = label2),
              vjust = 1.5, hjust = 0, size = 4.5, inherit.aes = FALSE) +
    theme_bw() +
    theme(
        strip.text = element_text(face = "bold"),
        legend.position = "right",
        axis.title.y = element_text(size = 15),
        axis.text.y = element_text(size = 15),
        axis.title.x = element_text(size = 15),
        axis.text.x = element_text(size = 15)
    ) +
    labs(
        y = "Geographic Distance (km)",
        x = expression(F[ST])
    ) + scale_color_manual(
  values = funcols,
  labels = function(x) stringr::str_wrap(x, width = 25)
) + guides(color=guide_legend(ncol=1))
p1
dev.off()
```

## Supplementary Figure S4

```{r, fig.height=10, fig.width=15}
mtclines = xx %>% mutate(Narrow_Term = if_else(is.na(Narrow_Term), "Non-MT", Narrow_Term)) %>% group_by(label2, Narrow_Term) %>% summarise(Fst = mean(Fst), distance = mean(distance)) %>% filter(!label2 %in% c("West vs West", "East vs East"))
mtclines$Narrow_Term = factor(mtclines$Narrow_Term, levels = c(unique(mtclines$Narrow_Term)[!grepl("Non-MT", unique(mtclines$Narrow_Term))], "Non-MT"))

West_vs_Selfe_terms = mtclines[,c(1,2,3)] %>% pivot_wider(names_from = label2, values_from = Fst) %>% filter(`West vs Selfe`*1.1 < `South Canterbury vs Selfe`) %>% filter(Narrow_Term != "kinetoplast") %>% dplyr::select(Narrow_Term) %>% as.list()
mtclines = mtclines %>% data.frame() %>% mutate(color = if_else(Narrow_Term %in% West_vs_Selfe_terms$Narrow_Term, Narrow_Term, ""))
mtclines = mtclines[mtclines$color != "",]
mtclines$color = factor(mtclines$color, levels =c(unique(mtclines$color)[-1],unique(mtclines$color)[1]))


line_labels <- mtclines %>% data.frame() %>%
    select(distance, label2) %>%
    distinct()

line_segments <- mtclines %>%
    arrange(Narrow_Term, distance) %>%
    group_by(Narrow_Term) %>%  
    mutate(
        next_distance = lead(distance),
        next_Fst = lead(Fst)
    ) %>%
    filter(!is.na(next_distance), (next_distance - distance) <= 50) %>%
    ungroup()
pdf(paste0(working_dir,"/../Figures/mtcategories_clines.narrow.pdf"), width = 9, height = 7)
px = data.frame("cat" = c("N-MT genes", "N-MT in selected GO terms"), "nb" = c(length(unique(xx[!is.na(xx$Narrow_Term),]$gene_name)), length(unique(xx[xx$Narrow_Term %in% West_vs_Selfe_terms$Narrow_Term,]$gene_name)))) %>% ggplot(aes(x="", y = nb, fill = cat)) + geom_col() + theme_bw() + theme(legend.position = "", axis.title.y = element_text(size = 15),
        axis.text.y = element_text(size = 15),
        axis.title.x = element_text(size = 15),
        axis.text.x = element_text(size = 15)) + labs(x="", y = "Number of genes") + scale_fill_manual(values = c("grey", "red"))
p2 = ggplot(mtclines[mtclines$color != "",], aes(x = Fst, y = distance)) +
    # Draw only short lines between points
    geom_segment(data = line_segments,
                 aes(x = Fst, xend = next_Fst,
                     y = distance, yend = next_distance,
                     color = color),
                 linewidth = 0.8,
                 inherit.aes = FALSE) +
    geom_point(aes(color = color), size = 2) +
    geom_hline(yintercept = unique(mtclines$distance), 
               linetype = "dashed", color = "black", alpha = 0.5) +
    geom_text(data = line_labels,
              aes(x = 0.065, y = distance, label = label2),
              vjust = 1.5, hjust = 0, size = 4, inherit.aes = FALSE) +
    theme_bw() +
    theme(
        strip.text = element_text(face = "bold"),
        legend.position = "right", legend.text = element_text(size = 11),
        axis.title.y = element_text(size = 15),
        axis.text.y = element_text(size = 15),
        axis.title.x = element_text(size = 15),
        axis.text.x = element_text(size = 15)
    ) +
    labs(
        y = "Geographic Distance (km)",
        x = expression(F[ST]), colour = "GO and KEGG terms"
    ) + scale_fill_manual(
        values = Cb64k,
        labels = function(x) str_wrap(x, width = 50)
    ) + scale_color_manual(
        values = Cb64k,
        labels = function(x) str_wrap(x, width = 50)
    ) + guides(fill=guide_legend(ncol=1), color=guide_legend(ncol=1))
px + p2 + plot_layout(widths = c(0.1,0.9)) + ggtitle("West less divergent from Selfe")
dev.off()
```

## Figure 4E

Manhattan plot of the mitochondrial functions for East vs Selfe

```{r, fig.height=5, fig.width=15}
xxx = xxx %>% mutate(Function2_color = case_when(Function2 == "Mitochondrial" & color == "not_outlier"~ "Mitochondrial not_outlier", Function2 == "Mitochondrial" & color == "final_outlier" ~ Category, .default = "rest")) 
xxx$Function2_color = str_remove(xxx$Function2_color, "GO_")
xxx$Function2_color = factor(xxx$Function2_color, levels = c("Mitochondrial not_outlier", unique(xxx$Function2_color)[c(-1,-2)], "rest"))
xxx$label = factor(xxx$label, levels = c("Alexandrina_Mapourika", "Alexandrina_Paringa", "Mapourika_Middleton", "Middleton_Paringa", "Alexandrina_Selfe", "Middleton_Selfe", "Paringa_Selfe", "Mapourika_Selfe", "Alexandrina_Middleton","Mapourika_Paringa"))
xxx = xxx[xxx$label %in% c("Alexandrina_Selfe", "Middleton_Selfe", "Paringa_Selfe", "Mapourika_Selfe"),]

Cb64k <- c(funcols[c(-12)], setNames(c("grey", "black"), c("rest", "Mitochondrial not_outlier")))


pdf(paste0(working_dir,"/../Figures/Manhattan_Mitochondrial.pdf"),height = 8, width = 10)
ggplot(xxx,aes(window,Fst)) + geom_point(data = xxx[xxx$Function2_color == "rest",], aes(window,Fst), color = "grey") + geom_point(data = xxx[xxx$Function2_color == "Mitochondrial not_outlier",], aes(window,Fst), color = "black") + geom_point(data = xxx[!(xxx$Function2_color %in% c("Mitochondrial not_outlier", "rest")),], aes(window,Fst, color = Function2_color), size = 3) + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) + facet_wrap(~label, ncol = 1) + scale_color_manual(values = funcols) + labs(color = "N-MT category", y = F[ST] ~ "value", x = "genomic window") + theme(axis.title.y = element_text(size = 15), strip.text = element_text(size = 15), axis.title.x =  element_text(size = 15), axis.text.y = element_text(size = 15))
dev.off()
```

## Figure 4A

Venn - how many overlap with bottom up analysis

```{r, fig.height=10, fig.width=15}

list_of_outliers = NULL
for (i in unique(for_plots$label)){
  tb = for_plots[for_plots$label == i & for_plots$color == "final_outlier",]
  tb = tb$window
  list_of_outliers[[i]] = tb
}

West_East_outliers = setdiff(setdiff(intersect(intersect(intersect(list_of_outliers$Alexandrina_Paringa, list_of_outliers$Alexandrina_Mapourika), list_of_outliers$Mapourika_Middleton), list_of_outliers$Middleton_Paringa), list_of_outliers$Alexandrina_Middleton), list_of_outliers$Mapourika_Paringa)

Incoming_to_Selfe_East= setdiff(setdiff(intersect(intersect(West_East_outliers, list_of_outliers$Mapourika_Selfe), list_of_outliers$Paringa_Selfe), list_of_outliers$Alexandrina_Selfe), list_of_outliers$Middleton_Selfe)
Incoming_to_Selfe_East=Incoming_to_Selfe_East[!duplicated(Incoming_to_Selfe_East)]
  
Incoming_to_Selfe_West= setdiff(setdiff(intersect(intersect(West_East_outliers, list_of_outliers$Middleton_Selfe), list_of_outliers$Alexandrina_Selfe), list_of_outliers$Mapourika_Selfe),list_of_outliers$Paringa_Selfe)
Incoming_to_Selfe_West=Incoming_to_Selfe_West[!duplicated(Incoming_to_Selfe_West)]

BDMI = setdiff(setdiff(setdiff(setdiff(West_East_outliers, list_of_outliers$Alexandrina_Selfe), list_of_outliers$Mapourika_Selfe), list_of_outliers$Paringa_Selfe), list_of_outliers$Middleton_Selfe)

obj = for_plots %>% 
  mutate(category = case_when(window %in% Incoming_to_Selfe_East ~ "Introgressing from South Canterbury", 
                           window %in% Incoming_to_Selfe_West ~ "Introgressing from West",
                           window %in% BDMI ~ "Introgressing from both",
                           .default = "Rest")) 
Incoming_to_Selfe_East_genes = setdiff(unique(obj[obj$category == "Introgressing from South Canterbury",]$gene_name), "")
Incoming_to_Selfe_West_genes = setdiff(unique(obj[obj$category == "Introgressing from West",]$gene_name), "")
BDMI_genes = setdiff(unique(obj[obj$category == "Introgressing from both",]$gene_name), "")

Mitochondrial=unique(unique(xx[xx$Category != "Rest",]$gene_name))

pdf(paste0(working_dir,"/../Figures/GGvenn_mito_WestvsEast_overlap.pdf"), height = 6, width = 9)
ggVennDiagram(list("Introgressing from both" = BDMI_genes, "Introgressing from South Canterbury"= Incoming_to_Selfe_East_genes, "Introgressing from West" = Incoming_to_Selfe_West_genes, "Mitochondrial" = Mitochondrial), set_color = c("orange", "plum", "darkcyan", "black"), label_color = "white", label_size = 5) + scale_fill_gradient2(midpoint = 150, low = "skyblue4", high = "pink4")  + scale_x_continuous(expand = expansion(mult = .4))
dev.off()
```

what if you subsample widows at random - how many will map to mitochondrial genes - MC permutation

```{r orths, fig.height=6, fig.width=6}

# Parameters
N <- length(unique(for_plots$gene_name))                # Total number of genes
x <- length(unique(bind_rows(mtcandidates)$Gene))                # Number of mitochondrial genes
n <- length(Incoming_to_Selfe_East_genes)                 # Number of outliers
y <-  length(intersect(unique(bind_rows(mtcandidates)$Gene), Incoming_to_Selfe_East_genes))                 # Observed number of mitochondrial outliers

# Monte Carlo simulation
set.seed(42)  # For reproducibility
num_simulations <- 10000
results <- replicate(num_simulations, {
  sampled_genes <- sample(unique(for_plots$gene_name), n)  # Randomly sample outlier genes
  sum(sampled_genes %in% bind_rows(mtcandidates)$Gene)  # Count mitochondrial genes in the sample
})


p_value <- mean(results >= y)

cat("Observed mitochondrial outliers:", y, "\n")
cat("P-value (MC permutation test):", p_value, "\n")
```

## Figure 4C

```{r, fig.height=10, fig.width=15}

pdf(paste0(working_dir,"/../Figures/GGvenn_mito_overlap.pdf"), height = 3, width = 5)
cats =list("South Canterbury vs Selfe" = unique(xx[(xx$label2 == "South Canterbury vs Selfe") & (xx$color == "final_outlier") & (xx$Function2 == "Mitochondrial"),]$window), "West vs Selfe" = unique(xx[(xx$label2 == "West vs Selfe") & (xx$color == "final_outlier") & (xx$Function2 == "Mitochondrial"),]$window), "West vs South Canterbury" = unique(xx[(xx$label2 == "West vs South Canterbury") & (xx$color == "final_outlier") & (xx$Function2 == "Mitochondrial"),]$window))
ggVennDiagram(cats, label_color = "white") + scale_x_continuous(expand = expansion(mult = .4)) + ggtitle("Diverget genomic windows within N-MT genes") 
dev.off()
```
