---
title:  "Fst sliding window analysis - divergence between West and South Canterbury"
author: "Natalia Zajac"
output:
  html_document: 
  highlight: pygments
  theme: sand
  code_folding: hide
  toc: yes
  toc_depth: 4
  editor_options: 
  chunk_output_type: inline
---

Started on `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`

```{r setup, include=FALSE}
library(poolfstat)
library(pheatmap)
library(dplyr)
library(cowplot)
library(patchwork)
library(readr)
library(stringr)
library(tidyverse)
library(textshaping)
library(ggVennDiagram)
library(ComplexHeatmap)
library(data.table)
library(ezRun)
library(scales)
library(ggpubr)
library(ggrepel)
library(enrichplot)
library(GO.db)
library(enrichR)
library(clusterProfiler)
library(GSEABase)
library(tidyverse)
working_dir=getwd()

##Load the dataframe of Fst per window (together with nucleotide diversity per window, results from Fisher's exact test and results from BayPass)
Fst_grendalf = read_delim(paste0(working_dir,"/../FST/Popoolation2/Sliding_Fst_per_Lake/Fst_sliding.Grendalf.Pi.FET.C2.txt"))

df = read_delim(paste0(working_dir,"/../test.sync")) # just for colnames
poolnames = str_remove(sapply(str_split(colnames(df)[c(-1:-3)], "\\."), .subset, 2), "B-")
metadata = data.frame(pool = poolnames , lake = sapply(str_split(poolnames, "_"), .subset, 1)) %>% column_to_rownames("pool") %>% mutate(lake = if_else(lake == "Alex", "Alexandrina", lake))
colors = list(lake = setNames(c("orange", "cornflowerblue","gold", "purple", "darkgreen"), unique(metadata$lake)))

## Load gene annotation and merge with the previous dataframe - annotation created with Genome_Assembly_Organization/GCA_013407085.1.R
geneAnno = read_delim(paste0(working_dir,"/../Atriophallophorus_winterbourni/EBI/GCA_013407085.1/Annotation/Genes/genes_annotation_byGene.txt"))

genes = data.table(geneAnno[,c(6,8,9)])
colnames(genes) = c("Chr", "start", "end")
setkey(genes, Chr, start, end)

snps = Fst_grendalf[,c(1,2,3)] %>% unique() 
snps = data.table(snps)
setkey(snps, chrom, start, end)
obj = foverlaps(snps, genes, by.x = c("chrom", "start", "end"), 
                by.y = c("Chr", "start", "end"), type = "within", nomatch = 0)
obj = merge(data.frame(obj), geneAnno[,c(3,6,8,9)], by.x = c("chrom", "start", "end"), by.y = c("seqid", "start", "end"), all.x = TRUE)
Fst_grendalf = merge(data.frame(Fst_grendalf), obj %>% dplyr::select(-2,-3) , by.x = c("chrom", "start", "end"), by.y = c("chrom", "i.start", "i.end"), all.x = TRUE) %>% replace(is.na(.), "")
Fst_grendalf$window = paste0(Fst_grendalf$chrom, "-", Fst_grendalf$start, "-", Fst_grendalf$end)
Fst_grendalf$window_length = Fst_grendalf$end - Fst_grendalf$start
Fst_grendalf = Fst_grendalf[Fst_grendalf$window_length > 0,]

##Create Gene Ontology Annotation df list
term2gene = NULL
term2gene[['BP']] = geneAnno[,c(3,12)] %>% filter(!is.na(`GO BP`)) %>%
    separate_rows(`GO BP`, sep = "; ") %>%
    dplyr::rename(Term = `GO BP`, Gene = gene_name) %>% dplyr::select(c(Term,Gene))
term2gene[['MF']] = geneAnno[,c(3,13)] %>% filter(!is.na(`GO MF`)) %>%
    separate_rows(`GO MF`, sep = "; ") %>%
    dplyr::rename(Term = `GO MF`, Gene = gene_name) %>% dplyr::select(c(Term,Gene))
term2gene[['CC']] = geneAnno[,c(3,14)] %>% filter(!is.na(`GO CC`)) %>%
    separate_rows(`GO CC`, sep = "; ") %>%
    dplyr::rename(Term = `GO CC`, Gene = gene_name) %>% dplyr::select(c(Term,Gene))

term2name= NULL
term2name[["BP"]] = data.frame(Term = term2gene$BP$Term, Name = sapply(term2gene$BP$Term, function(x){Term(x)[[1]]}))
term2name[["MF"]] = data.frame(Term = term2gene$MF$Term, Name = sapply(term2gene$MF$Term, function(x){Term(x)[[1]]}))
term2name[["CC"]] = data.frame(Term = term2gene$CC$Term, Name = sapply(term2gene$CC$Term, function(x){Term(x)[[1]]}))

#Annotate Gos with GOSSlims
myIds <- term2gene$BP$Term
myCollection <- GOCollection(myIds)
fl <- "../Atriophallophorus_winterbourni/EBI/GCA_013407085.1/GO_files/goslim.obo" 
slim <- getOBOCollection(fl)
slimdf <- goSlim(myCollection, slim, ontology = "BP")
GO.db::GOBPOFFSPRING
mappedIds <-
  function(df, collection, OFFSPRING)
  {
    map <- as.list(OFFSPRING[rownames(df)])
    mapped <- lapply(map, intersect, ids(collection))
    df[["go_terms"]] <- vapply(unname(mapped), paste, collapse = ";", character(1L))
    df
  }
slimdf <- mappedIds(slimdf, myCollection, GOBPOFFSPRING)
slimdf = slimdf %>% rownames_to_column("GOslim") %>% separate_longer_delim(go_terms, delim = ";")
slimdf = merge(slimdf, term2gene$BP, by.x = "go_terms", by.y = "Term")
slimdf = slimdf[,c(1,2,5,6)]
slimdfBP = slimdf[!duplicated(slimdf),] %>% mutate(GO = "BP")


myIds <- term2gene$MF$Term
myCollection <- GOCollection(myIds)
fl <- "../Atriophallophorus_winterbourni/EBI/GCA_013407085.1/GO_files/goslim.obo" 
slim <- getOBOCollection(fl)
slimdf <- goSlim(myCollection, slim, ontology = "MF")
GO.db::GOMFOFFSPRING
mappedIds <-
  function(df, collection, OFFSPRING)
  {
    map <- as.list(OFFSPRING[rownames(df)])
    mapped <- lapply(map, intersect, ids(collection))
    df[["go_terms"]] <- vapply(unname(mapped), paste, collapse = ";", character(1L))
    df
  }
slimdf <- mappedIds(slimdf, myCollection, GOMFOFFSPRING)
slimdf = slimdf %>% rownames_to_column("GOslim") %>% separate_longer_delim(go_terms, delim = ";")
slimdf = merge(slimdf, term2gene$MF, by.x = "go_terms", by.y = "Term")
slimdf = slimdf[,c(1,2,5,6)]
slimdfMF = slimdf[!duplicated(slimdf),] %>% mutate(GO = "MF")

myIds <- term2gene$CC$Term
myCollection <- GOCollection(myIds)
fl <- "../Atriophallophorus_winterbourni/EBI/GCA_013407085.1/GO_files/goslim.obo" 
slim <- getOBOCollection(fl)
slimdf <- goSlim(myCollection, slim, ontology = "CC")
GO.db::GOCCOFFSPRING
mappedIds <-
  function(df, collection, OFFSPRING)
  {
    map <- as.list(OFFSPRING[rownames(df)])
    mapped <- lapply(map, intersect, ids(collection))
    df[["go_terms"]] <- vapply(unname(mapped), paste, collapse = ";", character(1L))
    df
  }
slimdf <- mappedIds(slimdf, myCollection, GOCCOFFSPRING)
slimdf = slimdf %>% rownames_to_column("GOslim") %>% separate_longer_delim(go_terms, delim = ";")
slimdf = merge(slimdf, term2gene$CC, by.x = "go_terms", by.y = "Term")
slimdf = slimdf[,c(1,2,5,6)]
slimdfCC = slimdf[!duplicated(slimdf),] %>% mutate(GO = "CC")
slimdf = rbind(slimdfBP, slimdfMF, slimdfCC)

## OR read the precomputed df in because it takes a while to run the above
slimdf = read.delim("ogslimdf.txt")

## Load information content per GO
IC = read.delim("../Atriophallophorus_winterbourni/EBI/GCA_013407085.1/GO_files/Atrio_GO_IC", header = F, sep = " ")

#Add KO annotation
kegg = read.delim("../Gene2Kegg_Awinterbourni.txt",header = F, sep = "\t")
kegg = kegg[-1,c(1,2,3)]
term2gene$Kegg = kegg[kegg$V2 != "",c(2,1)]
colnames(term2gene$Kegg) = c("KO","Gene")
term2gene$Kegg$Gene = str_remove(term2gene$Kegg$Gene, "-mRNA-1")
term2name$Kegg = kegg[kegg$V2 != "",c(2,3)]
colnames(term2name$Kegg) = c("KO", "Name")


#Prepare minor allele freq data
rc = read.delim(paste0(working_dir,"/../FST/Popoolation2/combined.filtered.nomasked_rc"))
rc = rc %>% 
  mutate(across(starts_with("maa"), ~ as.numeric(sapply(str_split(., "/"), .subset,1))/as.numeric(sapply(str_split(., "/"), .subset,2)))) %>% 
  mutate(across(starts_with("mia"), ~ as.numeric(sapply(str_split(., "/"), .subset,1))/as.numeric(sapply(str_split(., "/"), .subset,2))))
rc = rc[,c(1:2,26:37,39:41)]
colnames(rc) = c("chrom", "snp", rownames(metadata)[-13])

#Calculate average allele frequency per each window
windows = Fst_grendalf[,c(1,2,3)]
windows = unique(windows)
windows = data.table(windows)
setkey(windows, chrom, start, end)
snps = rc[,c(1,2)]
snps = unique(snps)
snps = data.table(snps)
snps[, Pos2 := snp]
snps$snp = as.numeric(snps$snp)
snps$Pos2 = as.numeric(snps$Pos2)
setkey(snps, chrom, snp, Pos2)
obj = foverlaps(snps, windows, by.x = c("chrom", "snp", "Pos2"), by.y = c("chrom", "start", "end"), type = "within", nomatch = 0)
rc = merge(rc, data.frame(obj)[,-5], by = c("chrom", "snp"))
rc = reshape2::melt(rc, id.vars = c("chrom", "snp", "start", "end")) %>% dplyr::mutate(variable = sapply(str_split(variable, "_"), .subset, 1)) %>% group_by(variable, chrom, start, end) %>% dplyr::summarise(mia = mean(value))
rc = rc %>% mutate(variable = if_else(variable == "Alex", "Alexandrina", variable))

#Calculate dxy per window
dxy_per_window = NULL
for (i in unique(Fst_grendalf$label)[c(2:5,7:11,15)]){
  lake1 = sapply(str_split(i, "_"), .subset, 1)
  lake2 = sapply(str_split(i, "_"), .subset, 2)
  lake1df = rc[rc$variable == lake1,-1]
  colnames(lake1df) = c("chrom", "start", "end", lake1)
  lake2df = rc[rc$variable == lake2,-1]
  colnames(lake2df) = c("chrom", "start", "end", lake2)
  df = merge(lake1df, lake2df, by = c("chrom", "start", "end"))
  df[["value"]] <- df[[lake1]] * (1 - df[[lake2]]) + df[[lake2]] * (1 - df[[lake1]])
  df = df[,c(1,2,3,6)]
  df$variable = i
  dxy_per_window = rbind(dxy_per_window, df)
}
dxy_per_window$window = paste0(dxy_per_window$chrom, "-", dxy_per_window$start, "-", dxy_per_window$end)

## Compute average per window of each two-lake comparison
for_plots = NULL
for (i in unique(Fst_grendalf$label)[c(2:5,7:11,15)]){
  tb = Fst_grendalf[Fst_grendalf$label == i,]
  x = list("Grendalf" = tb[tb$MeanGrendalfFst_4sd == "outlier",]$window, 
           "C2" = tb[tb$C2std_Sig_SNPs > 0,]$window, 
           "FetMean" = tb[tb$FET_SigMean_SNPs > 0,]$window, 
           "FetMedian" = tb[tb$FET_SigMedian_SNPs > 0,]$window)
  tb$color = if_else(tb$window %in% Reduce(intersect, x), "final_outlier", "not_outlier")
  for_plots = rbind(for_plots, tb)
}
```


# {.tabset}

This analysis will be bottom up - finding differentiated windows between the West Lakes and South Canterbury Lakes and observing their pattern of divergence in Lake Selfe, then identifying genes associated with these windows and then performing GO enrichment analysis on these genes. Eventually the goal is to identify whether these functions could be implicated with mitonuclear discordance. 

## Analysis of outlier SNPs {.tabset}

### All comparisons together. 

```{r, fig.height=18, fig.width=16}

ggplot(for_plots, aes(window, Fst, color = color)) + geom_point() + scale_color_manual(values = c("red", "black")) + theme(axis.text.x = element_blank()) + labs(x = "Position", y = "Average Fst") + facet_wrap(~label, ncol = 1)
```

### Average Fst vs distance

```{r, fig.height=5, fig.width=7}
distance = paste0(working_dir,"/../tLCP_1400.txt")
distance = read.delim(distance, sep = " ")
distance = distance[rownames(distance) %in% unique(metadata$lake),colnames(distance) %in% unique(metadata$lake)] %>% rownames_to_column() %>% melt()
distance = distance[!duplicated(distance),]
distance$label = paste0(distance$rowname, "_", distance$variable)

sp <- ggscatter(for_plots %>% group_by(label) %>% summarise(meanFst = mean(Fst)) %>% left_join(distance, by = "label"), x = "value", y = "meanFst",
                add = "reg.line",  
                add.params = list(color = "blue", fill = "lightgray"), 
                conf.int = TRUE 
) + geom_label_repel(aes(label = label)) + labs(x = "Distance", y = "Mean sliding window Fst")
# Add correlation coefficient
sp + stat_cor(method = "pearson", label.x = 50, label.y = 0.075)
```

### Number of genes

```{r, fig.height=5, fig.width=7}
for_plots %>% group_by(gene_name, color, label) %>% dplyr::summarise(count = n_distinct(window)) %>% filter(gene_name != "") %>% filter(count >1) %>% group_by(color, label) %>% dplyr::summarise(gene_count = n_distinct(gene_name)) %>% filter(color == "final_outlier") %>% ggplot(aes(gene_count, factor(label, levels = rev(c("Alexandrina_Middleton", "Mapourika_Paringa", "Alexandrina_Paringa", "Alexandrina_Mapourika", "Mapourika_Middleton", "Middleton_Paringa", "Alexandrina_Selfe", "Middleton_Selfe", "Mapourika_Selfe", "Paringa_Selfe"))))) + geom_col()

test = for_plots %>% group_by(gene_name, color, label) %>% dplyr::summarise(count = n_distinct(window)) %>% filter(gene_name != "") %>% filter(count >1) %>% filter(color == "final_outlier")
list_of_diff_genes = sapply(split( test[,c(1,3)] , f = test$label), function(x){return(x$gene_name)})

ggVennDiagram(list_of_diff_genes)
```

### length of windows

```{r, fig.height=5, fig.width=7}
windows = for_plots[,c(1,2,3,4,5,18,19,20,21)]
windows$window_length = windows$end - windows$start
windows$window_length_bins = ezCut(windows$window_length, breaks = seq(500, 5000, 1000))

windows[windows$color == "final_outlier",] %>% group_by(label, window_length_bins) %>% dplyr::summarise(count = n_distinct(window)) %>% ggplot(aes(count, factor(label, levels = rev(c("Alexandrina_Middleton", "Mapourika_Paringa", "Alexandrina_Paringa", "Alexandrina_Mapourika", "Mapourika_Middleton", "Middleton_Paringa", "Alexandrina_Selfe", "Middleton_Selfe", "Mapourika_Selfe", "Paringa_Selfe"))), fill = window_length_bins)) + geom_col(position = "fill") + labs(x = "Proportion", y = "", fill = "Window Length") + scale_fill_manual(values = brewPalette(n = 15))
```

### Sinlge bar showing all outlier sharing

First check how many genomic regions belong to which category

```{r, fig.height=5, fig.width=7}
temp = unique(for_plots[,c(4,19,20)]) %>% group_by(window, color) %>% dplyr::summarise(values = paste(label, collapse = ","))
name = paste(unique(for_plots$label), collapse = ",")
x1 = temp[temp$values == name & temp$color == "final_outlier",]
x2 = temp[temp$values == name & temp$color == "not_outlier",]
x3 = temp[temp$values != name & temp$color == "not_outlier",]
x4 = temp[temp$values != name & temp$color == "final_outlier",]
windows = list("all_non_outliers" = x1$window, "all_outliers" = x2$window, "not_all_outliers" = x4$window, "not_all_not_outliers" = x3$window)

ggVennDiagram(windows, label_color = "white")
```

## Figure 3A

```{r, fig.height=5, fig.width=11}
library(ggpp)
temp = unique(for_plots[,c(4,19,20)]) %>% group_by(window, color) %>% dplyr::summarise(values = paste(label, collapse = ","))
name = paste(unique(for_plots$label), collapse = ",")
x1 = temp[temp$values == name & temp$color == "final_outlier",] %>% mutate(category1 = "All comparisons differentiated", West_vs_East = "")
x2 = temp[temp$values == name & temp$color == "not_outlier",] %>% mutate(category1 = "All comparisons undifferentiated", West_vs_East = "")
x3 = temp[temp$values != name & temp$color == "not_outlier",] %>% mutate(category1 = "All comparisons", West_vs_East = "")
x4 = temp[temp$values != name & temp$color == "final_outlier",]

required_west_vs_east <- c("Alexandrina_Paringa", "Alexandrina_Mapourika", "Mapourika_Middleton", "Middleton_Paringa")

x4 = x4 %>%
    rowwise() %>% 
    mutate(
        # Split the `values` into individual lake pairs
        value_list = str_split(values, ","),
        # Check if all required elements are present in the value_list
        category1 = dplyr::case_when(
            all(required_west_vs_east %in% unlist(value_list)) ~ "West vs South Canterbury",
            values == "Alexandrina_Selfe,Mapourika_Selfe,Middleton_Selfe,Paringa_Selfe" ~ "Selfe vs all",
            values == "Alexandrina_Paringa,Alexandrina_Mapourika,Alexandrina_Selfe,Alexandrina_Middleton" ~ "Alexandrina vs all",
            values == "Alexandrina_Middleton,Mapourika_Middleton,Middleton_Selfe,Middleton_Paringa" ~ "Middleton vs all",
            values == "Alexandrina_Mapourika,Mapourika_Middleton,Mapourika_Paringa,Mapourika_Selfe" ~ "Mapourika vs all",
            values == "Alexandrina_Paringa,Mapourika_Paringa,Paringa_Selfe,Middleton_Paringa" ~ "Paringa vs all",
            values == "Alexandrina_Middleton" ~ "within South Canterbury",
            values == "Mapourika_Paringa" ~ "within the West",
            values %in% c("Alexandrina_Mapourika", "Alexandrina_Paringa", "Alexandrina_Selfe", 
                          "Mapourika_Middleton", "Mapourika_Selfe", "Middleton_Paringa", 
                          "Middleton_Selfe", "Paringa_Selfe") ~ "between only two lakes",
            values == "Alexandrina_Selfe,Middleton_Selfe" ~ "Selfe vs South Canterbury",
            values == "Mapourika_Selfe,Paringa_Selfe" ~ "Selfe vs West",
            TRUE ~ "Other"
        ), West_vs_East = dplyr::case_when(
      values == "Alexandrina_Paringa,Alexandrina_Mapourika,Mapourika_Middleton,Middleton_Paringa" ~ "Introgressing to Selfe from both",
      values == "Alexandrina_Paringa,Alexandrina_Mapourika,Mapourika_Middleton,Mapourika_Selfe,Paringa_Selfe,Middleton_Paringa" ~ "Introgressing to Selfe from South Canterbury",
      values == "Alexandrina_Paringa,Alexandrina_Mapourika,Alexandrina_Selfe,Mapourika_Middleton,Middleton_Selfe,Middleton_Paringa" ~ "Introgressing to Selfe from West",
      values %in%  c("Alexandrina_Paringa,Alexandrina_Mapourika,Alexandrina_Selfe,Mapourika_Middleton,Middleton_Selfe,Paringa_Selfe,Middleton_Paringa", "Alexandrina_Paringa,Alexandrina_Mapourika,Alexandrina_Selfe,Mapourika_Middleton,Mapourika_Paringa,Middleton_Selfe,Paringa_Selfe,Middleton_Paringa") ~ "Introgressing into Selfe from Mapourika",
      values %in%  c("Alexandrina_Paringa,Alexandrina_Mapourika,Alexandrina_Selfe,Mapourika_Middleton,Mapourika_Selfe,Middleton_Selfe,Middleton_Paringa", "Alexandrina_Paringa,Alexandrina_Mapourika,Alexandrina_Selfe,Mapourika_Middleton,Mapourika_Paringa,Mapourika_Selfe,Middleton_Selfe,Middleton_Paringa") ~ "Introgressing into Selfe from Paringa", 
      values %in% c("Alexandrina_Paringa,Alexandrina_Mapourika,Alexandrina_Selfe,Mapourika_Middleton,Mapourika_Selfe,Paringa_Selfe,Middleton_Paringa","Alexandrina_Paringa,Alexandrina_Mapourika,Alexandrina_Selfe,Alexandrina_Middleton,Mapourika_Middleton,Mapourika_Selfe,Paringa_Selfe,Middleton_Paringa") ~ "Introgressing into Selfe from Middleton",
      values %in% c("Alexandrina_Paringa,Alexandrina_Mapourika,Mapourika_Middleton,Mapourika_Selfe,Middleton_Selfe,Paringa_Selfe,Middleton_Paringa","Alexandrina_Paringa,Alexandrina_Mapourika,Alexandrina_Middleton,Mapourika_Middleton,Mapourika_Selfe,Middleton_Selfe,Paringa_Selfe,Middleton_Paringa") ~ "Introgressing into Selfe from Alexandrina",
      .default = ""))
x4 = x4[,c(-2,-4)]

temp2 = merge(unique(for_plots[,c(4,5,19,20)]), x4, by = c("window"))
temp3 = merge(unique(for_plots[,c(4,5,19,20)]), x1, by = c("window","color"))
temp4 = merge(unique(for_plots[,c(4,5,19,20)]), x2, by = c("window","color"))

temp3 = rbind(temp2,temp3,temp4)

temp3$category1 = factor(temp3$category1, levels = c("All comparisons undifferentiated", "All comparisons differentiated","Alexandrina vs all", "Middleton vs all", "Selfe vs all", "Mapourika vs all", "Paringa vs all", "between only two lakes", "Selfe vs South Canterbury", "Selfe vs West", "within South Canterbury",  "within the West", "West vs South Canterbury", "Other"))
temp3$West_vs_East = factor(temp3$West_vs_East, levels = c("Introgressing into Selfe from Alexandrina", "Introgressing into Selfe from Middleton", "Introgressing into Selfe from Paringa", "Introgressing into Selfe from Mapourika", "Introgressing to Selfe from South Canterbury", "Introgressing to Selfe from West", "Introgressing to Selfe from both", ""))

temp3$window_length = as.numeric(sapply(str_split(temp3$window, "-"), .subset,3)) - as.numeric(sapply(str_split(temp3$window, "-"), .subset,2))
```


```{r, fig.height=5, fig.width=11}
p1 = temp3 %>% mutate(category2 = if_else(category1 == "All comparisons undifferentiated", "Undifferentiated", "Differentiated in at least\none pairwise comparison")) %>% distinct(category2, window, window_length) %>%
    group_by(category2) %>% 
    dplyr::summarise(window_count = n_distinct(window), length = sum(window_length)) %>% mutate(perc = (length * 100)/sum(length)) %>% 
    mutate(perc = if_else(perc > 10, paste0(as.character(round(perc,0)),"%"), "")) %>% 
    ggplot(aes(x = "", y = window_count, fill = category2)) + geom_bar(stat = "identity", width = 1) + 
    coord_polar(theta = "y", start = 1.75) +
    labs(y = "", x = "", fill = "Comparison") +
    theme_bw() +
    theme(axis.ticks = element_blank(), 
          axis.text = element_blank(), 
          panel.grid = element_blank(), 
          panel.border = element_blank(),legend.text = element_text(size = 11), legend.position = "bottom", legend.title.position = "top") + scale_fill_manual(values = c("salmon","azure4")) +
    geom_label_repel(aes(label = perc),position = position_stacknudge(vjust = 0.5, x = 0.8), max.overlaps = 30, size = 9, show.legend = FALSE,color = "white")  + guides(fill = guide_legend(ncol = 1)) + labs(fill = "Genomic regions")


p2 = temp3 %>% filter(category1 != "All comparisons differentiated" & category1 != "All comparisons undifferentiated") %>% mutate(category2 = if_else(category1 == "West vs South Canterbury", "in West vs South Canterbury\ncomparisons", "Other")) %>% distinct(category2, window, window_length) %>% 
    group_by(category2) %>% 
    dplyr::summarise(window_count = n_distinct(window), length = sum(window_length)) %>% mutate(perc = (length*100)/173173889) %>% mutate(perc = if_else(perc > 0.5, paste0(as.character(round(perc,1)),"%"), "")) %>% 
    ggplot(aes(x = "", y = window_count, fill = category2)) + geom_bar(stat = "identity", width = 1) + 
    coord_polar(theta = "y", start = 1.9) +
    labs(y = "", x = "", fill = "Comparison") +
    theme_bw() +
    theme(axis.ticks = element_blank(), 
          axis.text = element_blank(), 
          panel.grid = element_blank(), 
          panel.border = element_blank(), legend.position = "bottom", legend.text = element_text(size = 11), legend.title.position = "top") + scale_fill_manual(values = c("#0086ED","salmon")) +
    geom_label_repel(aes(label = perc), position = position_stacknudge(vjust = 0.5, x = 0.8), max.overlaps = 30, size = 9, show.legend = FALSE, color = "white") + guides(fill = guide_legend(ncol = 1)) + labs(fill = "Divergent genomic regions")

p3 = temp3 %>% filter(category1 == "West vs South Canterbury") %>% mutate(West_vs_East = if_else(West_vs_East %in% c("", "Introgressing into Selfe from Alexandrina", "Introgressing into Selfe from Mapourika", "Introgressing into Selfe from Middleton", "Introgressing into Selfe from Paringa"), "Other", West_vs_East)) %>% distinct(West_vs_East, window, window_length) %>% group_by(West_vs_East) %>% dplyr::summarise(window_count = n_distinct(window), length = sum(window_length))  %>% mutate(perc = (length*100)/173173889) %>% mutate(perc = if_else(perc > 0.05, paste0(as.character(round(perc,2)),"%"), "")) %>% ggplot(aes(x = "", y = window_count, fill = West_vs_East)) + geom_bar(stat = "identity", width = 1,color = "#0086ED") +
    coord_polar(theta = "y") +
    labs(y = "", x = "", fill = "West vs East") +
    theme_bw() + scale_fill_manual(values = c("darkorange", "plum", "darkcyan", "#A6CEE3FF")) +
    geom_label_repel(aes(label = perc), position = position_stacknudge(vjust = 0.5, x = 0.8), max.overlaps = 30, size = 9, show.legend = FALSE, color = "white") + labs(fill = "Divergent between West vs South Canterbury") + guides(fill = guide_legend(ncol = 1)) +
    theme(axis.ticks = element_blank(), 
          axis.text = element_blank(), 
          panel.grid = element_blank(), 
          panel.border = element_blank(), legend.text = element_text(size = 11), legend.position = "bottom", legend.title.position = "top" )

p = p1 + p2 + p3 

pdf(paste0(working_dir,"/../Figures/Whole_genome_subdivided.v2.pdf"), width = 14, height = 6)
p
dev.off()

```

## Figure 3B

```{r, fig.height=5, fig.width=11}

for_fst = temp3 %>% mutate(
  label2 = case_when(label %in% c("Alexandrina_Paringa", "Alexandrina_Mapourika", "Middleton_Paringa", "Mapourika_Middleton") ~ "West vs South Canterbury", 
                     label %in% c("Alexandrina_Selfe", "Middleton_Selfe") ~ "South Canterbury vs Selfe",  label %in% c("Mapourika_Selfe", "Paringa_Selfe") ~ "West vs Selfe", label == "Alexandrina_Middleton" ~ "within South Canterbury",  
                     label == "Mapourika_Paringa" ~ "within the West"), 
  Comparison = case_when(category1 == "All comparisons undifferentiated" ~ "Undifferentiated", category1 != "All comparisons undifferentiated" & category1 != "West vs South Canterbury" ~ "Differentiated in at least\none pairwise comparison", category1 == "West vs South Canterbury" & !(West_vs_East %in% c("Introgressing to Selfe from South Canterbury", "Introgressing to Selfe from West", "Introgressing to Selfe from both")) ~ "West vs South Canterbury", West_vs_East %in% c("Introgressing to Selfe from South Canterbury", "Introgressing to Selfe from West", "Introgressing to Selfe from both") ~ paste0(category1,"\n",West_vs_East))) 
for_fst$Comparison = factor(for_fst$Comparison, levels = c("Undifferentiated", "Differentiated in at least\none pairwise comparison", "West vs South Canterbury", "West vs South Canterbury\nIntrogressing to Selfe from both", "West vs South Canterbury\nIntrogressing to Selfe from South Canterbury", "West vs South Canterbury\nIntrogressing to Selfe from West"))


p4 = for_fst %>% 
    ggplot(aes(label2, Fst, fill = Comparison)) + 
    geom_violin()  + 
    scale_fill_manual(values = c("azure4", "salmon" , "#A6CEE3FF", "darkorange", "plum","darkcyan")) + theme(axis.text.x = element_text(size = 13, angle = 45, hjust = 1, vjust = 1), plot.margin = margin(10, 10, 10, 50),axis.title.y = element_text(size = 18), axis.text.y = element_text(size = 15),legend.text = element_text(size = 11), legend.key.size = unit(1, "cm"), legend.key = element_rect(colour = NA, fill = NA)) + labs(x = "") + guides(fill=guide_legend(ncol=1)) + ylab(expression(F[ST])) + labs(fill = "Genomic regions")

pdf(paste0(working_dir,"/../Figures/Whole_genome_subdivided_FST.2.pdf"), width = 10, height = 5)
p4
dev.off()

```

## Figure 3D

```{r, fig.height=8, fig.width=10}
xxx = for_plots %>% mutate(east = case_when(
  window %in% for_fst[for_fst$Comparison == "West vs South Canterbury\nIntrogressing to Selfe from South Canterbury",]$window ~ "Introgressing from SC", 
  window %in% for_fst[for_fst$Comparison == "West vs South Canterbury\nIntrogressing to Selfe from both",]$window ~ "Introgressing from both", 
  window %in% for_fst[for_fst$Comparison == "West vs South Canterbury\nIntrogressing to Selfe from West",]$window ~ "Introgressing from West",
  .default = "Rest")) %>% arrange(chrom,start, end)

xxx$east = factor(xxx$east, levels = c("Introgressing from both", "Introgressing from SC", "Introgressing from West","Rest"))
xxx = xxx[xxx$label %in% c("Alexandrina_Selfe", "Middleton_Selfe", "Paringa_Selfe", "Mapourika_Selfe"),]
xxx$label = factor(xxx$label, levels = c("Alexandrina_Selfe", "Middleton_Selfe", "Paringa_Selfe", "Mapourika_Selfe"))

pdf(paste0(working_dir,"/../Figures/Manhattan_WestvsEast.v2.pdf"), height = 9, width = 9)

ggplot(xxx,aes(window,Fst)) + 
  geom_point(data = xxx[xxx$east == "Rest",], aes(window,Fst), color = "grey") + 
  geom_point(data = xxx[xxx$east == "Introgressing from both",], aes(window,Fst, color = east), size = 2) + 
  geom_point(data = xxx[xxx$east =="Introgressing from SC",], aes(window,Fst, color = east), size = 2) + 
  geom_point(data = xxx[xxx$east == "Introgressing from West",], aes(window,Fst, color = east), size = 2) + 
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) + 
  facet_wrap(~label, ncol = 1) + scale_color_manual(values = c("darkorange","plum","darkcyan")) + 
  labs(color = "West vs East into Selfe", y = F[ST] ~ "value", x = "genomic window") + 
  theme(axis.title.y = element_text(size = 18), axis.text.y = element_text(size = 13), strip.text = element_text(size = 17), axis.title.x =  element_text(size = 15))

dev.off()
```

## Figure 3C

```{r, include=FALSE}

West_East_outliers = setdiff(setdiff(intersect(intersect(intersect(list_of_outliers$Alexandrina_Paringa, list_of_outliers$Alexandrina_Mapourika), list_of_outliers$Mapourika_Middleton), list_of_outliers$Middleton_Paringa), list_of_outliers$Alexandrina_Middleton), list_of_outliers$Mapourika_Paringa)

Incoming_to_Selfe_East= setdiff(setdiff(intersect(intersect(West_East_outliers, list_of_outliers$Mapourika_Selfe), list_of_outliers$Paringa_Selfe), list_of_outliers$Alexandrina_Selfe), list_of_outliers$Middleton_Selfe)
Incoming_to_Selfe_East=Incoming_to_Selfe_East[!duplicated(Incoming_to_Selfe_East)]
  
Incoming_to_Selfe_West= setdiff(setdiff(intersect(intersect(West_East_outliers, list_of_outliers$Middleton_Selfe), list_of_outliers$Alexandrina_Selfe), list_of_outliers$Mapourika_Selfe),list_of_outliers$Paringa_Selfe)
Incoming_to_Selfe_West=Incoming_to_Selfe_West[!duplicated(Incoming_to_Selfe_West)]

BDMI = setdiff(setdiff(setdiff(setdiff(West_East_outliers, list_of_outliers$Alexandrina_Selfe), list_of_outliers$Mapourika_Selfe), list_of_outliers$Paringa_Selfe), list_of_outliers$Middleton_Selfe)

colors2 = setNames(c("plum", "darkcyan", "darkorange", "darkgrey"), c("Introgressing from South Canterbury", "Introgressing from West", "Introgressing from both", "Rest"))
```

```{r, fig.height=5, fig.width=11}

temp = for_plots[,c(1,2,3, 10,11,12,13,14,19)] %>% 
  mutate(color = case_when(window %in% Incoming_to_Selfe_East ~ "Introgressing from South Canterbury", 
                           window %in% Incoming_to_Selfe_West ~ "Introgressing from West",
                           window %in% BDMI ~ "Introgressing from both",
                           .default = "Rest")) %>% 
  reshape2::melt(id.vars = c("chrom", "start", "end", "color", "window"))

temp = temp[temp$color != "Rest",] %>% mutate(variable = str_remove(variable, "_ThetaPi"), 
                     variable2 = case_when(variable %in% c("Alexandrina", "Middleton") ~ "South Canterbury",
                                           variable %in% c("Mapourika", "Paringa") ~ "West", 
                                           .default = "Selfe"))
lev = combn(unique(temp$color), 2, simplify = FALSE) 
p1 = ggplot(temp,aes(color,
           value, fill = color)) +
    geom_violin() +
    theme(legend.position = "top") +
    stat_summary(fun = median, geom = "point", size = 2, color = "red",
                 position = position_dodge(width = 0.9)) +
    labs(x = "", y = expression(paste("Nucleotide diversity ", pi)), fill = F[ST] ~ "outlier")  +
    stat_compare_means(aes(group = color),
                       comparisons = lev,  # defined above
                       method = "wilcox.test",        # or "t.test"
                       label = "p.signif", p.adjust.method = "BH") + scale_fill_manual(values = colors2) + theme(legend.position = "top", axis.text.x = element_blank(),axis.text.y = element_text(size = 12), axis.title.y = element_text(size = 18), strip.text = element_text(size = 18), legend.text = element_text(size = 12)) + facet_grid(~factor(variable2, levels = c("South Canterbury", "Selfe", "West")))
pdf("../Figures/Nucleotide_diversity_WestvsEast.pdf", height = 5, width = 9)
p1
dev.off()
```

## Supplementary Figure S3

```{r, fig.width=13, fig.height=7}

obj = for_plots %>% mutate(category = case_when(window %in% Incoming_to_Selfe_East ~ "Introgressing from SC", window %in% Incoming_to_Selfe_West ~ "Introgressing from West", window %in% BDMI ~ "Introgressing from both", .default = ""))

tmp = merge(
  obj[obj$category != "",] %>% mutate(in_gene = if_else(gene_name != "", "in gene", "not in gene")), 
  merge(
    obj %>% 
      filter(category != "") %>% 
      group_by(category, gene_name, chrom) %>% 
      dplyr::summarise(windows1 = n_distinct(window)) %>% 
      filter(gene_name != ""), 
    obj %>% 
      group_by(gene_name, chrom) %>% 
      dplyr::summarise(windows2 = n_distinct(window)) %>% 
      filter(gene_name != ""), 
    by = c("gene_name", "chrom")) %>% 
    mutate(ratio = windows1/windows2), by = c("gene_name", "chrom","category"), all = TRUE)

tmp$window_length = tmp$end - tmp$start
tmp$window_length_bins = ezCut(tmp$window_length, breaks = seq(500, 10000, 1000))

tmp$category = factor(tmp$category, levels = c("Introgressing from SC", "Introgressing from West", "Introgressing from both"))

p1 = tmp %>% group_by(category, in_gene, window_length_bins) %>% dplyr::summarise(count = n_distinct(window)) %>% ggplot(aes(count, str_wrap(category,12), fill = window_length_bins)) + geom_col(position = "stack") + labs(x = "Window count", y = "", fill = "Window Length") + scale_fill_manual(values = brewPalette(n = 15)) + facet_grid(~in_gene) + theme(strip.text = element_text(size = 15), axis.text.y = element_text(size = 12))

p2 = ggplot(unique(tmp[,c(1,2,3,26)]), aes(category,ratio)) + ggbeeswarm::geom_quasirandom() + labs(x = "", y = "Fraction of gene divergence")

Incoming_to_Selfe_East_genes = setdiff(unique(obj[obj$category == "Introgressing from SC",]$gene_name), "")
Incoming_to_Selfe_West_genes = setdiff(unique(obj[obj$category == "Introgressing from West",]$gene_name), "")
BDMI_genes = setdiff(unique(obj[obj$category == "Introgressing from both",]$gene_name), "")

list_of_genes_of_interest = list("Introgressing from SC"= Incoming_to_Selfe_East_genes, "Introgressing from West" = Incoming_to_Selfe_West_genes, "Introgressing from both" = BDMI_genes)

p3 = ggVennDiagram(list_of_genes_of_interest, set_color = unname(colors2)[c(1,2,3)], label_color  = "white") + scale_x_continuous(expand = expansion(mult = .4))

pdf("../Figures/Characterization_of_WestEast_genomicregions.pdf", height = 7, width = 12)
p = p2 + p3
p1/p
dev.off()
```


## GO enrichment analysis {.tabset}

```{r}

list_of_genes_of_interest$`Introgressing from both` = setdiff(setdiff(list_of_genes_of_interest$`Introgressing from both`, list_of_genes_of_interest$`Introgressing from East`), list_of_genes_of_interest$`Introgressing from West`)

BP_enrich = NULL
MF_enrich = NULL
CC_enrich = NULL

for (i in names(list_of_genes_of_interest)){
  df = enricher(list_of_genes_of_interest[[i]], 
               pvalueCutoff = 0.05, 
               pAdjustMethod = "BH", 
               geneAnno$gene_name, 
               minGSSize = 1, 
               maxGSSize = 500, 
               qvalueCutoff = 0.2, 
               TERM2GENE = term2gene$BP, 
               TERM2NAME = term2name$BP)
  df = df@result %>% mutate(GO = "BP") %>% 
    mutate("category" = i) %>% 
    mutate(GeneRatio = as.numeric(sapply(str_split(GeneRatio, "/"), .subset, 1))/as.numeric(sapply(str_split(GeneRatio, "/"), .subset, 2))) %>% 
    filter(qvalue < 0.05)
  BP_enrich = rbind(BP_enrich, df)
  dfm = enricher(list_of_genes_of_interest[[i]], 
               pvalueCutoff = 0.05, 
               pAdjustMethod = "BH", 
               geneAnno$gene_name, 
               minGSSize = 1, 
               maxGSSize = 500, 
               qvalueCutoff = 0.2, 
               TERM2GENE = term2gene$MF, 
               TERM2NAME = term2name$MF)
  dfm = dfm@result %>% mutate(GO = "MF") %>% 
    mutate("category" = i) %>% 
    mutate(GeneRatio = as.numeric(sapply(str_split(GeneRatio, "/"), .subset, 1))/as.numeric(sapply(str_split(GeneRatio, "/"), .subset, 2))) %>% 
    filter(qvalue < 0.05)
  MF_enrich = rbind(MF_enrich, dfm)
  dfc = enricher(list_of_genes_of_interest[[i]], 
               pvalueCutoff = 0.05, 
               pAdjustMethod = "BH", 
               geneAnno$gene_name, 
               minGSSize = 1, 
               maxGSSize = 500, 
               qvalueCutoff = 0.2, 
               TERM2GENE = term2gene$CC, 
               TERM2NAME = term2name$CC)
  dfc = dfc@result %>% mutate(GO = "CC") %>% 
    mutate("category" = i) %>% 
    mutate(GeneRatio = as.numeric(sapply(str_split(GeneRatio, "/"), .subset, 1))/as.numeric(sapply(str_split(GeneRatio, "/"), .subset, 2))) %>% 
    filter(qvalue < 0.05)
  CC_enrich = rbind(CC_enrich, dfc)
}

```

### Figure 3E

```{r, fig.height=10, fig.width=18}
combined = rbind(BP_enrich, MF_enrich, CC_enrich) 
combined = combined  %>% separate_longer_delim(geneID, delim = "/")
combined = merge(combined, IC, by.x = "ID", by.y = "V1", all.x = TRUE)


subset_combined = NULL
for (i in unique(combined$geneID)){
  for (x in unique(combined$category)){
    for (c in unique(combined$GO)){
      if (nrow(combined[which(combined$geneID == i & combined$category == x & combined$GO == c),]) > 0){
      txm = combined[which(combined$geneID == i & combined$category == x & combined$GO == c),]
      txm = txm[txm$V2 == min(txm$V2),]
      subset_combined = rbind(subset_combined, txm)
      }
    }
  }
}


subset_combined = merge(subset_combined, slimdf[,c(1,2,5,6,7,8)], by.x = c("ID","geneID", "GO"), by.y = c("go_terms","Gene", "GO"), all.x = TRUE)
desc = setNames(subset_combined$Description, subset_combined$Term)
desc = desc[!duplicated(desc)]
desc = desc[!is.na(names(desc))]

rbind(combined[combined$Count > 1 & combined$category %in% c("Differentiated West-East", "Differentiated Selfe-West"),], combined[combined$category %in% c("Differentiated Selfe-East"),])
# subset_combined = subset_combined[!subset_combined$Description %in% c("hydrolase activity, acting on carbon-nitrogen (but not peptide) bonds, in nitriles"),]
subset_combined = subset_combined %>% mutate(Term = if_else(is.na(Term), "Not annotated", Term))
subset_combined = subset_combined %>% mutate(category = case_when(category == "Introgressing from West"~ "from West", category == "Introgressing from East"~ "Introgressing from SC", .default = category))
subset_combined$category = factor(subset_combined$category, levels = c("Introgressing from SC","from West", "Introgressing from both"))
subset_combined$Term = factor(subset_combined$Term, levels = c("Not annotated", unique(subset_combined$Term)[!grepl("Not annotated", unique(subset_combined$Term))]))

p1 = subset_combined %>% 
  ggplot(aes(GeneRatio,str_wrap(Description, 35), size = Count,color = pvalue)) + 
  geom_point() + 
  facet_grid(category~GO, scales = "free",space = "free_y") + labs(y = "") + theme_bw() +
  theme(axis.text.x = element_text(size = 8), strip.text = element_text(size = 18)) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm"), legend.position = "right",axis.text.y = element_blank(), legend.text = element_text(size = 11)) + scale_size_continuous(breaks = c(1, 2, 3)) + xlim(0,0.25)

p2 = ggplot(subset_combined) +
    geom_col(aes(y = str_wrap(Description, 35), x = 1, fill = Term), position = "fill") + facet_grid(category~., scales = "free",space = "free_y") + 
    theme_bw()  + 
    theme(axis.text.x = element_blank(), 
          axis.title.x = element_blank(), 
          axis.ticks.x = element_blank(), 
          panel.background = element_blank(), 
          plot.background = element_blank(), strip.text = element_blank(), axis.title.y = element_blank(), plot.margin = unit(c(0, 0, 0, 0), "cm"), axis.text.y = element_text(size = 15,margin = margin(t = 10, b = 10)),legend.text = element_text(size = 15)) + scale_fill_manual(values = c("white",brewPalette(n = 21))) + guides(fill=guide_legend(ncol=1))

p = p2 + p1 +
    plot_layout(widths = c(.05, 1.1), ncol = 2, guides = "collect") & theme(legend.position = 'right') 

# p3 = combined  %>% separate_longer_delim(geneID, delim = "/") %>% 
#   group_by(GO, category) %>% 
#   summarise(count = n_distinct(geneID)) %>% 
#   ggplot(aes(GO,count)) + 
#   geom_col() + 
#   labs(x = "GO category", y = "Number of genes") + 
#   facet_grid(category~GO, scales = "free_x") + 
#   theme(axis.text.x = element_blank()) 
# 
# plot_grid(p, p3, ncol = 2, rel_widths = c(3, 1))

#write_delim(data.frame(GOslim = unique(combined$Term)[!is.na(unique(combined$GOslim))]), "../Sex_specific_genes/GO_terms_WEdiff_of_interest.txt")
pdf("../Figures/GO_results.pdf", height = 12, width = 14)
p
dev.off()
```

### Additional related plots - emapplots

```{r, fig.height=8, fig.width=10}
edo = enricher(list_of_genes_of_interest$`Introgressing from East`, 
               pvalueCutoff = 0.05, 
               pAdjustMethod = "BH", 
               geneAnno$gene_name, 
               minGSSize = 1, 
               maxGSSize = 500, 
               qvalueCutoff = 0.2, 
               TERM2GENE = term2gene$MF, 
               TERM2NAME = term2name$MF)

edox = pairwise_termsim(edo)
pdf("../Figures/emapplot_MF_fromEast.pdf", height=8, width=10)
emapplot(edox, cex_label_category = 0.7,  layout = "nicely")
dev.off()
```

```{r, fig.height=8, fig.width=10}
edo = enricher(list_of_genes_of_interest$`Introgressing from West`, 
               pvalueCutoff = 0.05, 
               pAdjustMethod = "BH", 
               geneAnno$gene_name, 
               minGSSize = 1, 
               maxGSSize = 500, 
               qvalueCutoff = 0.2, 
               TERM2GENE = term2gene$MF, 
               TERM2NAME = term2name$MF)

edox = pairwise_termsim(edo)
pdf("../Figures/emapplot_MF_fromWest.pdf", height=5, width=5)
emapplot(edox, cex_label_category = 0.7,  layout = "nicely")
dev.off()
```


```{r, fig.height=8, fig.width=10}
edo = enricher(list_of_genes_of_interest$`Introgressing from both`, 
               pvalueCutoff = 0.05, 
               pAdjustMethod = "BH", 
               geneAnno$gene_name, 
               minGSSize = 1, 
               maxGSSize = 500, 
               qvalueCutoff = 0.2, 
               TERM2GENE = term2gene$MF, 
               TERM2NAME = term2name$MF)

edox = pairwise_termsim(edo)
pdf("../Figures/emapplot_MF_fromBoth.pdf", height=5, width=7)
emapplot(edox, cex_label_category = 0.7,  layout = "nicely")
dev.off()
```

