---
title:  "Whole genome pairwise FST - characterization of population structure"
author: "Natalia Zajac"
output:
  html_document: 
  highlight: pygments
theme: sand
code_folding: hide
toc: yes
toc_depth: 4
editor_options: 
  chunk_output_type: inline
---

Started on `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`


```{r setup, include=FALSE}
library(poolfstat)
library(pheatmap)
library(dplyr)
library(cowplot)
library(patchwork)
library(readr)
library(stringr)
library(tidyverse)
library(textshaping)
library(ggpubr)
library(ggrepel)

working_dir = getwd()

## Load baypass utils for plotting omega covariance matrix
source("../BayPass/baypass_utils.R")

df = read_delim(paste0(working_dir,"/../test.sync")) # just for colnames
poolnames = str_remove(sapply(str_split(colnames(df)[c(-1:-3)], "\\."), .subset, 2), "B-")

#Nuclear genome sync file loaded as pooldata using poolfstat
pooldata = popsync2pooldata(
  sync.file = paste0(working_dir,"/../combined.filtered.nomasked.sync"),
  poolsizes = c(rep(24,16)),
  poolnames = poolnames,
  nthreads = 1, min.maf = 0.0001
)
pooldata = pooldata.subset(pooldata, pool.index = c(1:12,14:16))
nuc = (data.frame(pooldata@refallele.readcount)* 100)/data.frame(pooldata@readcoverage)
colnames(nuc) = pooldata@poolnames
nuc = cbind(data.frame(pooldata@snp.info), nuc)


#Mitochondrial genome sync data loaded as pooldata using poolfstat
pooldatamt = popsync2pooldata(
  sync.file = paste0(working_dir,"/../all_samples.mit.sync.filtered.sync"),
  poolsizes = c(rep(12,15)),
  poolnames = poolnames[-13],
  noindel = TRUE,
  nlines.per.readblock = 1e+06,
  nthreads = 1, min.maf = 0.0001
)

#More stringent noise filtering (VAF < 5% across all)
mtcov= data.frame(pooldatamt@readcoverage)
mtcov = cbind(data.frame(pooldatamt@snp.info), mtcov)
tokeep = mtcov %>%
    filter(if_all(5:19, ~ . < 6000))
pooldatamt = poolfstat::pooldata.subset(pooldatamt, snp.index = as.numeric(rownames(tokeep)))
mito = (data.frame(pooldatamt@refallele.readcount)* 100)/data.frame(pooldatamt@readcoverage)
colnames(mito) = pooldatamt@poolnames
mito = cbind(data.frame(pooldatamt@snp.info), mito)
to_filter = rownames(mito[rowSums(mito[, 5:19] >= 92) == 15 | rowSums(mito[, 5:19] <= 8) == 15, ])
pooldatamt = poolfstat::pooldata.subset(pooldatamt, snp.index = as.numeric(rownames(mito)[!rownames(mito) %in% to_filter]))
mito = mito[!rownames(mito) %in% to_filter,]

#Filter and save sync file for computing FST
#df = read.delim("../all_samples.mit.sync.filtered.sync", header = F)
#write_delim(df[!df$V2 %in% mito[rowSums(mito[, 5:19] >= 95) == 15 | rowSums(mito[, 5:19] <= 5) == 15, c(2)],], "../all_samples.mit.sync.filtered.VAF5.sync")

#Create metadata
metadata = data.frame(pool = poolnames , lake = sapply(str_split(poolnames, "_"), .subset, 1)) %>% column_to_rownames("pool") %>% mutate(lake = if_else(lake == "Alex", "Alexandrina", lake))

colors = list(lake = setNames(c("orange", "cornflowerblue","gold", "purple", "darkgreen"), unique(metadata$lake)))
geneAnno = read_delim("../Atriophallophorus_winterbourni/EBI/GCA_013407085.1/Annotation/Genes/genes_annotation_byGene.txt")

##Read in distance matrices
dst = NULL
for (i in list.files(paste0(working_dir,"/../Distance_matrices/"))){
  distance = str_remove(str_remove(i, "tLCP_"), ".txt")
  x = read.delim(file.path("../Distance_matrices", i), sep = " ")
  x = x[rownames(x) %in% unique(metadata$lake),colnames(x) %in% unique(metadata$lake)]
  dst[[distance]] = x
}

#Read in pairwise genome wide Fst matrices computed with Grenedalf
Fst_gren_Hudson = read_delim(paste0(working_dir,"/../FST/PairwiseFst.Hudsonfst-matrix.csv"), delim = ",") %>% column_to_rownames("fst")
colnames(Fst_gren_Hudson) = rownames(metadata)
rownames(Fst_gren_Hudson) = rownames(metadata)
Fst_gren_Hudson = Fst_gren_Hudson[-13,-13]

Fst_gren_Hudson_mt = read_delim("../FST/PairwiseFst.Hudson.mtfst-matrix.csv", delim = ",") %>% column_to_rownames("fst")
colnames(Fst_gren_Hudson_mt) = rownames(metadata[-13, , drop = FALSE])
rownames(Fst_gren_Hudson_mt) = rownames(metadata[-13, , drop = FALSE])

#Read in sliding window Fst
Fst_grendalf = read_delim(paste0(working_dir,"/../FST/Popoolation2/Sliding_Fst_per_Lake/Fst_sliding.Grendalf.Pi.FET.C2.txt"))

#Read in omega matrices from BayPass
omega_cov_matrix = read.table(paste0(working_dir,"/../BayPass/Alexandrina_Mapourika_mat_omega.out"), header = F) 
omega_cov_matrix=cov2cor(as.matrix(omega_cov_matrix))
colnames(omega_cov_matrix) = rownames(metadata)[-13]
rownames(omega_cov_matrix) = rownames(metadata)[-13]

omega_cov_matrix_mt = read.table(paste0(working_dir,"/../BayPass/all_samples.mit.Alexandrina_Selfe_mat_omega.out"), header = F) #THis is interepreted with caution because mitochondrium is haploid and so not suitable for running Baypass
omega_cov_matrix_mt=cov2cor(as.matrix(omega_cov_matrix_mt))
colnames(omega_cov_matrix_mt) = rownames(metadata)[-13]
rownames(omega_cov_matrix_mt) = rownames(metadata)[-13]
```

## Effective Sample Sizes - initial testing - subsequently estimated with poolne_estim

```{r, fig.height=5, fig.width=10}
ess = data.frame(pooldata@readcoverage)
colnames(ess) = c("Alexandrina1", "Alexandrina2", "Alexandrina3", "Mapourika1", "Mapourika2", "Mapourika3", "Middleton1", "Middleton2", "Middleton3", "Paringa1", "Paringa2", "Paringa3", "Selfe1", "Selfe2","Selfe3")
colors_custom = list(lake = setNames(c(rep("orange", 3), rep("cornflowerblue",3),rep("gold",3), rep("purple",3), rep("darkgreen",3)), c("Alexandrina1", "Alexandrina2", "Alexandrina3", "Mapourika1", "Mapourika2", "Mapourika3", "Middleton1", "Middleton2", "Middleton3", "Paringa1", "Paringa2", "Paringa3", "Selfe1", "Selfe2","Selfe3")))
ess$cat = "read cov"
ess = reshape2::melt(ess, id.vars = "cat")
ess$value_mod = (ess$value * 24)/(ess$value + 24)
ggplot(ess, aes(value_mod, color = variable)) + geom_density(bw=1) + scale_color_manual(values = colors_custom$lake) + theme_bw() + labs(x = "Effective Sample Size", y = "Density", colour = "PoolNames") + theme(text = element_text(size = 15))
```

## Supplementary Figure 2

```{r, fig.height=5, fig.width=10}
pdf(paste0(working_dir, "../Figures/Mitochondrial_genome_reference_allele_percentage_238snps.pdf"), width = 7, height = 5)
reshape2::melt(mito, id.vars = c("Chromosome", "Position", "RefAllele" ,"AltAllele")) %>% mutate(variable = sapply(str_split(variable, "_"), .subset, 1)) %>% mutate(variable = if_else(variable == "Alex", paste0(variable, "andrina", seq(1,3,1)), paste0(variable, seq(1,3,1)))) %>% ggplot(aes(value, colour = variable)) + geom_density() + scale_colour_manual(values = c("orange","orange","orange", "cornflowerblue","cornflowerblue","cornflowerblue","gold","gold","gold", "purple","purple","purple", "darkgreen","darkgreen","darkgreen")) + theme_bw()
dev.off()

data_for_pheatmap = mito[, c(5:19)]
colnames(data_for_pheatmap) = c("Alexandrina1", "Alexandrina2", "Alexandrina3", "Mapourika1", "Mapourika2", "Mapourika3", "Middleton1", "Middleton2", "Middleton3", "Paringa1", "Paringa2", "Paringa3", "Selfe1", "Selfe2", "Selfe3")

pdf(paste0(working_dir,"../Figures/Mitochondrial_Variants_Allele_Frequenciespheatmap.pdf"), width = 5, height = 7)
pheatmap(data_for_pheatmap, show_rownames = F)
dev.off()
```

## Pairwise FST - different visualisations

```{r, fig.height=5, fig.width=10}
pairwisefst = compute.pairwiseFST(
  pooldata,
  method = "Identity",
  min.cov.per.pool = -1,
  max.cov.per.pool = 1e+06,
  min.indgeno.per.pop = -1,
  min.maf = -1,
  output.snp.values = FALSE,
  nsnp.per.bjack.block = 0,
  verbose = TRUE
)

pairwisefstmt = compute.pairwiseFST(
  pooldatamt,
  method = "Identity",
  min.cov.per.pool = -1,
  max.cov.per.pool = 1e+06,
  min.indgeno.per.pop = -1,
  min.maf = -1,
  output.snp.values = FALSE,
  nsnp.per.bjack.block = 0,
  verbose = TRUE
)

pheatmap(pairwisefst@PairwiseFSTmatrix, 
         annotation_col = metadata[-13, , drop = FALSE], 
         annotation_colors = colors, 
         show_colnames = F, name = "Fst")


vals <- pairwisefst@values
vals=vals %>% rownames_to_column() %>% separate(rowname, into = c("pool1", "pool2"), sep = ";")
df = vals %>% group_by(pool1,pool2) %>% summarise(Fst = median(`Fst Estimate`))
df = df %>% filter(pool1 != pool2)
df = df %>% mutate(pool1 = case_when(pool1 == "Alex_CAMP2ndWP" ~ "Alexandrina1", pool1 == "Alex_MW" ~ "Alexandrina2", pool1 == "Alex_NE" ~ "Alexandrina3", .default = pool1)) %>% mutate(pool2 = case_when(pool2 == "Alex_CAMP2ndWP" ~ "Alexandrina1", pool2 == "Alex_MW" ~ "Alexandrina2", pool2 == "Alex_NE" ~ "Alexandrina3", .default = pool2)) %>% mutate(pool1 = str_remove(pool1, "_Pool")) %>% mutate(pool2 = str_remove(pool2, "_Pool")) %>% mutate(pool2 = str_remove(pool2, "_Pool")) 
df[df$pool1 == "Mapourika_Otto3",]$pool1 = "Mapourika3"
df[df$pool2 == "Mapourika_Otto3",]$pool2 = "Mapourika3"
df[df$pool1 == "Mapourika_Mix2",]$pool1 = "Mapourika2"
df[df$pool2== "Mapourika_Mix2",]$pool2 = "Mapourika2"
df[df$pool1 == "Mapourika_Jetty1",]$pool1 = "Mapourika1"
df[df$pool2 == "Mapourika_Jetty1",]$pool2 = "Mapourika1"
df = rbind(df, df %>% rename(pool1=pool2, pool2 = pool1)) 
df$pool1 = factor(df$pool1, levels = c("Alexandrina1", "Alexandrina2", "Alexandrina3","Middleton1", "Middleton2","Middleton3", "Selfe1", "Selfe2","Selfe3", "Paringa1","Paringa2","Paringa3", "Mapourika1", "Mapourika2","Mapourika3"))
df$pool2 = factor(df$pool2, levels = c("Alexandrina1", "Alexandrina2", "Alexandrina3","Middleton1", "Middleton2","Middleton3", "Selfe1", "Selfe2","Selfe3", "Paringa1","Paringa2","Paringa3", "Mapourika1", "Mapourika2","Mapourika3"))
df$idx1 <- as.integer(df$pool1)
df$idx2 <- as.integer(df$pool2)
df <- df %>% filter(idx1 < idx2)
px = ggplot(df, aes(pool1, pool2, fill=Fst)) +
    geom_tile() +
    coord_fixed() +
    theme_minimal() + theme(text = element_text(size = 20), axis.text.x = element_blank()) + labs(x="", y = "")  + scale_fill_continuous(palette = "RdYlBu")


vals <- pairwisefstmt@values
vals=vals %>% rownames_to_column() %>% separate(rowname, into = c("pool1", "pool2"), sep = ";")
df = vals %>% group_by(pool1,pool2) %>% summarise(Fst = median(`Fst Estimate`))
df = df %>% filter(pool1 != pool2)
df = df %>% mutate(pool1 = case_when(pool1 == "Alex_CAMP2ndWP" ~ "Alexandrina1", pool1 == "Alex_MW" ~ "Alexandrina2", pool1 == "Alex_NE" ~ "Alexandrina3", .default = pool1)) %>% mutate(pool2 = case_when(pool2 == "Alex_CAMP2ndWP" ~ "Alexandrina1", pool2 == "Alex_MW" ~ "Alexandrina2", pool2 == "Alex_NE" ~ "Alexandrina3", .default = pool2)) %>% mutate(pool1 = str_remove(pool1, "_Pool")) %>% mutate(pool2 = str_remove(pool2, "_Pool")) %>% mutate(pool2 = str_remove(pool2, "_Pool")) 
df[df$pool1 == "Mapourika_Otto3",]$pool1 = "Mapourika3"
df[df$pool2 == "Mapourika_Otto3",]$pool2 = "Mapourika3"
df[df$pool1 == "Mapourika_Mix2",]$pool1 = "Mapourika2"
df[df$pool2== "Mapourika_Mix2",]$pool2 = "Mapourika2"
df[df$pool1 == "Mapourika_Jetty1",]$pool1 = "Mapourika1"
df[df$pool2 == "Mapourika_Jetty1",]$pool2 = "Mapourika1"
df = rbind(df, df %>% rename(pool1=pool2, pool2 = pool1)) 
df$pool1 = factor(df$pool1, levels = c("Alexandrina1", "Alexandrina2", "Alexandrina3","Middleton1", "Middleton2","Middleton3", "Selfe1", "Selfe2","Selfe3", "Paringa1","Paringa2","Paringa3", "Mapourika1", "Mapourika2","Mapourika3"))
df$pool2 = factor(df$pool2, levels = c("Alexandrina1", "Alexandrina2", "Alexandrina3","Middleton1", "Middleton2","Middleton3", "Selfe1", "Selfe2","Selfe3", "Paringa1","Paringa2","Paringa3", "Mapourika1", "Mapourika2","Mapourika3"))
df$idx1 <- as.integer(df$pool1)
df$idx2 <- as.integer(df$pool2)
df <- df %>% filter(idx1 > idx2)
pn = ggplot(df, aes(pool1, pool2, fill=Fst)) +
    geom_tile() +
    coord_fixed() +
    theme_minimal() + theme(text = element_text(size = 20), axis.text.x = element_blank()) + labs(x="", y = "")  + scale_fill_continuous(palette = "RdYlBu") +scale_y_discrete(position = "right")

```

## Congruence between FST and distance matrices - CADM - Figure 1F

```{r, fig.height=5, fig.width=10}
Fst = data.frame(pairwisefst@PairwiseFSTmatrix) %>% rownames_to_column("pool1") %>% pivot_longer(!pool1) %>% mutate(lake1 = sapply(str_split(pool1, "_"), .subset, 1), lake2 = sapply(str_split(name, "_"), .subset, 1)) %>% group_by(lake1, lake2) %>% mutate(lake1 = if_else(lake1 == "Alex", "Alexandrina", lake1), lake2 = if_else(lake2 == "Alex", "Alexandrina", lake2))  %>% dplyr::summarise(Fst = mean(value)) %>% pivot_wider(names_from = lake2, values_from = Fst) %>% column_to_rownames("lake1") %>% as.matrix() %>% replace(is.na(.), 0)

Fstmt = data.frame(pairwisefstmt@PairwiseFSTmatrix) %>% rownames_to_column("pool1") %>% pivot_longer(!pool1) %>% mutate(lake1 = sapply(str_split(pool1, "_"), .subset, 1), lake2 = sapply(str_split(name, "_"), .subset, 1)) %>% group_by(lake1, lake2) %>% mutate(lake1 = if_else(lake1 == "Alex", "Alexandrina", lake1), lake2 = if_else(lake2 == "Alex", "Alexandrina", lake2))  %>% dplyr::summarise(Fst = mean(value)) %>% pivot_wider(names_from = lake2, values_from = Fst) %>% column_to_rownames("lake1") %>% as.matrix() %>% replace(is.na(.), 0)

Fst_Hudson = data.frame(Fst_gren_Hudson) %>% rownames_to_column("pool1") %>% pivot_longer(!pool1) %>% mutate(lake1 = sapply(str_split(pool1, "_"), .subset, 1), lake2 = sapply(str_split(name, "_"), .subset, 1)) %>% group_by(lake1, lake2) %>% mutate(lake1 = if_else(lake1 == "Alex", "Alexandrina", lake1), lake2 = if_else(lake2 == "Alex", "Alexandrina", lake2)) %>% mutate(value = if_else(lake1 == lake2, 0, value)) %>% dplyr::summarise(Fst = mean(value)) %>% pivot_wider(names_from = lake2, values_from = Fst) %>% column_to_rownames("lake1") %>% as.matrix() %>% replace(is.na(.), 0)

Fst_Hudson_mt = data.frame(Fst_gren_Hudson_mt) %>% rownames_to_column("pool1") %>% pivot_longer(!pool1) %>% mutate(lake1 = sapply(str_split(pool1, "_"), .subset, 1), lake2 = sapply(str_split(name, "_"), .subset, 1)) %>% group_by(lake1, lake2) %>% mutate(lake1 = if_else(lake1 == "Alex", "Alexandrina", lake1), lake2 = if_else(lake2 == "Alex", "Alexandrina", lake2)) %>% mutate(value = if_else(lake1 == lake2, 0, value)) %>% dplyr::summarise(Fst = mean(value)) %>% pivot_wider(names_from = lake2, values_from = Fst) %>% column_to_rownames("lake1") %>% as.matrix() %>% replace(is.na(.), 0)

list_of_fst_matrices = list("IIS_Fst" = Fst, "IIS_Fst_mt" = Fstmt, "Hudson_Unbiased" = Fst_Hudson, "Hudson_Unbiased_mt" = Fst_Hudson_mt)

all_cong = NULL
for (x in names(list_of_fst_matrices)){
  n = list_of_fst_matrices[[x]]
  congruence = NULL
  for (i in names(dst)){
    cng = CADM.global(rbind(data.frame(n), data.frame(dst[[i]])[rownames(data.frame(n)), colnames(data.frame(n))]), 2, 5, nperm=99, make.sym=TRUE, weights=NULL, silent=FALSE)
    congruence[[i]] = cng$congruence_analysis
    }
  congruence = bind_rows(congruence, .id = "Sample")
  names = colnames(congruence)
  congruence = data.frame(congruence)
  rownames(congruence) = c("W", "Chi2", "Prob.perm")
  colnames(congruence) = names
  congruence = t(congruence) %>% data.frame() %>% rownames_to_column("Distance") %>% mutate(p_value = if_else(as.numeric(Prob.perm) > 0.05, "P-value > 0.05", "P-value < 0.05"), group = 1)
  all_cong[[x]] = congruence
}

pdf(paste0(working_dir,"/../Figures/CADM.pdf"), width = 7, height = 7)
bind_rows(all_cong, .id = "Matrix_type") %>% ggplot(aes(as.integer(Distance), W, shape = p_value, group = Matrix_type, linetype= Matrix_type)) + geom_point(size = 3) + geom_line() + labs(x = "Distance", y = str_wrap("Kendall's coefficient of concordance (CADM)", width = 50)) + scale_shape_manual(values = c(19,21)) + labs(linetype = expression(F[ST] ~ "measurement")) + scale_linetype_manual(values = c("longdash", "dotted", "twodash", "solid")) + theme(legend.position = "inside", legend.position.inside = c(0.2, 0.3), axis.text = element_text(size = 18), axis.title = element_text(size = 18), legend.text = element_text(size = 15)) + geom_vline(xintercept = 1400, color = "red", linetype = "dashed", size = 0.5)+ ggtitle("Congruence between genetic and geographic distance")
dev.off()
```


## Pariwise Fst vs the best distance estimation - Figure 1E

```{r, fig.height=5, fig.width=10}
Fst_sliding = Fst_grendalf %>% group_by(label) %>% summarise(sliding_window_Hudson_Fst = mean(Fst))

Fst_pairwise_Hudson = Fst_gren_Hudson %>% rownames_to_column("pool1") %>% reshape2::melt() %>% mutate(lake1 = sapply(str_split(pool1, "_"), .subset, 1), lake2 = sapply(str_split(variable, "_"), .subset, 1)) %>% group_by(lake1, lake2) %>% mutate(lake1 = if_else(lake1 == "Alex", "Alexandrina", lake1), lake2 = if_else(lake2 == "Alex", "Alexandrina", lake2))  %>% summarise(pairwise_Hudson_Fst = mean(value)) %>% mutate(label = paste0(lake1, "_", lake2)) %>% ungroup() %>% dplyr::select(label, pairwise_Hudson_Fst) %>% filter(label %in% unique(Fst_sliding$label))

Fst_pairwise_Hudson_mt = Fst_gren_Hudson_mt %>% rownames_to_column("pool1") %>% reshape2::melt() %>% mutate(lake1 = sapply(str_split(pool1, "_"), .subset, 1), lake2 = sapply(str_split(variable, "_"), .subset, 1)) %>% group_by(lake1, lake2) %>% mutate(lake1 = if_else(lake1 == "Alex", "Alexandrina", lake1), lake2 = if_else(lake2 == "Alex", "Alexandrina", lake2))  %>% summarise(pairwise_Hudson_Fst_mt = mean(value)) %>% mutate(label = paste0(lake1, "_", lake2)) %>% ungroup() %>% dplyr::select(label, pairwise_Hudson_Fst_mt) %>% filter(label %in% unique(Fst_sliding$label))

combined_Fst = Fst_sliding %>% left_join(Fst_pairwise_Hudson, by = "label") %>% left_join(Fst_pairwise_Hudson_mt, by = "label") %>% reshape2::melt(id.vars = "label") 

distance = paste0(working_dir,"/../tLCP_1400.txt")
distance = read.delim(distance, sep = " ")
distance = distance[rownames(distance) %in% unique(metadata$lake),colnames(distance) %in% unique(metadata$lake)] %>% rownames_to_column() %>% reshape2::melt()
distance = distance[!duplicated(distance),]
distance$label = paste0(distance$rowname, "_", distance$variable)
colnames(distance) = c("lake1", "lake2", "distance", "label")

df = combined_Fst %>% filter(variable != "sliding_window_Hudson_Fst") %>% left_join(distance, by = "label") %>% mutate(lab = if_else(grepl("Selfe",label), label, "")) %>% filter(sapply(str_split(label, "_"), .subset, 1) != sapply(str_split(label, "_"), .subset, 2))
df = df %>% mutate(variable = if_else(variable == "pairwise_Hudson_Fst", "Nuclear", "Mitochondrial"))
df$variable = factor(df$variable, levels = c("Nuclear", "Mitochondrial"))

ggscatter(df, x = "distance", y = "value", add = "reg.line", conf.int = TRUE) + facet_wrap(~variable, scales = "free_y") + stat_cor(method = "pearson", label.x = 0) + xlim(60, 360) + geom_point(aes(color = lab), size = 3) + scale_color_manual(values = c("black", "orange", "cornflowerblue", "gold", "purple")) + labs(x = "Distance (km) at 1400m flight altitude", y = F[ST] ~ "value") + theme_bw() + theme(legend.position = "", text = element_text(size = 20)) + stat_cor(method = "pearson", label.y.npc = 1) 

## Alternative versions

sp1 = ggscatter(combined_Fst %>% filter(variable != "sliding_window_Hudson_Fst") %>% left_join(distance, by = "label") %>% mutate(color = case_when(label %in% c("Middleton_Selfe", "Alexandrina_Selfe") ~ "group1", label %in% c("Mapourika_Selfe", "Paringa_Selfe") ~ "group2", .default = "group3")) %>% filter(variable == "pairwise_Hudson_Fst"), x = "distance", y = "value", add = "reg.line", conf.int = TRUE) + theme(legend.position = "") + guides(color = guide_legend(ncol = 2)) + geom_label_repel(aes(label = label)) + stat_cor(method = "pearson", label.x = 0) + geom_point(aes(color = color)) + scale_color_manual(values = c("red", "yellow", "black")) + labs(x = "Distance", y = "Fst value") + facet_wrap(~variable, scales = "free", ncol = 2)

sp2 = ggscatter(combined_Fst %>% filter(variable != "sliding_window_Hudson_Fst") %>% left_join(distance, by = "label") %>% mutate(color = case_when(label %in% c("Middleton_Selfe", "Alexandrina_Selfe") ~ "group1", label %in% c("Mapourika_Selfe", "Paringa_Selfe") ~ "group2", .default = "group3")) %>% filter(variable == "pairwise_Hudson_Fst_mt"), x = "distance", y = "value", add = "reg.line", conf.int = TRUE) + theme(legend.position = "") + guides(color = guide_legend(ncol = 2)) + geom_label_repel(aes(label = label)) + stat_cor(method = "pearson", label.x = 0, label.y = 1.2) + geom_point(aes(color = color)) + scale_color_manual(values = c("red", "yellow", "black")) + labs(x = "Distance", y = "Fst value") + facet_wrap(~variable, scales = "free", ncol = 2)

sp1 = ggscatter(combined_Fst[!combined_Fst$label %in% c("Selfe_Selfe", "Middleton_Middleton", "Paringa_Paringa", "Mapourika_Mapourika", "Alexandrina_Alexandrina"),] %>% filter(variable != "sliding_window_Hudson_Fst") %>% left_join(distance, by = "label") %>% mutate(color = case_when(label %in% c("Middleton_Selfe", "Alexandrina_Selfe") ~ "group1", label %in% c("Mapourika_Selfe", "Paringa_Selfe") ~ "group2", .default = "group3")) %>% filter(variable == "pairwise_Hudson_Fst"), x = "distance", y = "value", add = "reg.line", conf.int = TRUE) + theme(legend.position = "") + guides(color = guide_legend(ncol = 2)) + geom_label_repel(aes(label = label)) + stat_cor(method = "pearson", label.x = 0) + geom_point(aes(color = color)) + scale_color_manual(values = c("red", "yellow", "black")) + labs(x = "Distance (km) at 1400m maximum flight altitude", y = F[ST] ~ "value") 
sp1 = sp1 + ggtitle("Pairwise Hudson" ~ F[ST], subtitle = "nuclear data")

sp2 = ggscatter(combined_Fst[!combined_Fst$label %in% c("Selfe_Selfe", "Middleton_Middleton", "Paringa_Paringa", "Mapourika_Mapourika", "Alexandrina_Alexandrina"),] %>% filter(variable != "sliding_window_Hudson_Fst") %>% left_join(distance, by = "label") %>% mutate(color = case_when(label %in% c("Middleton_Selfe", "Alexandrina_Selfe") ~ "group1", label %in% c("Mapourika_Selfe", "Paringa_Selfe") ~ "group2", .default = "group3")) %>% filter(variable == "pairwise_Hudson_Fst_mt"), x = "distance", y = "value", add = "reg.line", conf.int = TRUE) + theme(legend.position = "") + guides(color = guide_legend(ncol = 2)) + geom_label_repel(aes(label = label)) + stat_cor(method = "pearson", label.x = 0, label.y = 1.2) + geom_point(aes(color = color)) + scale_color_manual(values = c("red", "yellow", "black")) + labs(x = "Distance (km) at 1400m maximum flight altitude", y = F[ST] ~ "value") 
sp2 = sp2 + ggtitle("Pairwise Hudson" ~ F[ST], subtitle = "mitochondrial data")

sp1 /sp2 
```

## Figure 1C

```{r, fig.height=5, fig.width=10}

colnames(Fst_gren_Hudson) = c("Alexandrina1", "Alexandrina2", "Alexandrina3", "Mapourika1", "Mapourika2", "Mapourika3", "Middleton1", "Middleton2", "Middleton3", "Paringa1", "Paringa2", "Paringa3", "Selfe1", "Selfe2", "Selfe3")
rownames(Fst_gren_Hudson) = c("Alexandrina1", "Alexandrina2", "Alexandrina3", "Mapourika1", "Mapourika2", "Mapourika3", "Middleton1", "Middleton2", "Middleton3", "Paringa1", "Paringa2", "Paringa3", "Selfe1", "Selfe2", "Selfe3")

pdf(paste0(working_dir,"/../Figures/Fst_plot_Nuc.pdf"), height = 3, width = 5)
p1 = ComplexHeatmap::pheatmap(as.matrix(Fst_gren_Hudson), annotation_col = metadata[-13, , drop = FALSE] , annotation_colors = colors, show_colnames = F, name = "Fst", breaks = seq(0.02, 0.1, length.out = 20), main = "Nuclear", fontsize = 10)
dev.off() 

colnames(Fst_gren_Hudson_mt) = c("Alexandrina1", "Alexandrina2", "Alexandrina3", "Mapourika1", "Mapourika2", "Mapourika3", "Middleton1", "Middleton2", "Middleton3", "Paringa1", "Paringa2", "Paringa3", "Selfe1", "Selfe2", "Selfe3")
rownames(Fst_gren_Hudson_mt) = c("Alexandrina1", "Alexandrina2", "Alexandrina3", "Mapourika1", "Mapourika2", "Mapourika3", "Middleton1", "Middleton2", "Middleton3", "Paringa1", "Paringa2", "Paringa3", "Selfe1", "Selfe2", "Selfe3")
pdf("../Figures/Fst_plot_Mt.pdf", height = 3, width = 5)
p2 = ComplexHeatmap::pheatmap(as.matrix(Fst_gren_Hudson_mt), annotation_col = metadata[-13, , drop = FALSE] , annotation_colors = colors, show_colnames = F, name = "Fst_mt", breaks = seq(0, 1, length.out = 25), main = "Mitochondrial")
dev.off() 
```

## PCA based on Reference Allele Frequency {.tabset}

### Nuclear  - Figure 1D (left hand side)

```{r, fig.height=5, fig.width=10}

rc = t(as.matrix(nuc[,c(5:19)]/100))
rc = rc[,colSums(rc) > 0]
rc = rc[,colSums(rc > 0) >= 3]
rc = rc[,colSums(rc < 1) >= 3] # choose SNPs where ref allele freq is greater than 0 or less than 1 in at least 3 pools
pc = prcomp(rc, center = TRUE, scale. = TRUE)
summ <- summary(pc)
summ
rownames(pc$x) = c("Alexandrina1", "Alexandrina2", "Alexandrina3", "Mapourika1", "Mapourika2", "Mapourika3", "Middleton1", "Middleton2", "Middleton3", "Paringa1", "Paringa2", "Paringa3", "Selfe1", "Selfe2", "Selfe3")
rownames(metadata) = c("Alexandrina1", "Alexandrina2", "Alexandrina3", "Mapourika1", "Mapourika2", "Mapourika3", "Middleton1", "Middleton2", "Middleton3", "Paringa1", "Paringa2", "Paringa3", "Paringa4", "Selfe1", "Selfe2", "Selfe3")

df =  pc$x %>% data.frame() %>% rownames_to_column("pool") %>% left_join(metadata[-13, , drop = FALSE] %>% rownames_to_column("pool"), by = "pool") %>% mutate(circle = if_else(lake %in% c("Alexandrina", "Middleton", "Selfe"), "South Canterbury and Lake Selfe", "West"))

p1 = df %>% ggplot(aes(PC1, PC2, color = lake)) + geom_point(size = 3) + theme_bw() + theme(legend.position = "") + scale_color_manual(values = colors$lake) + geom_label_repel(aes(label = pool)) + labs(x = "PC1 (11.6%)",y = "PC2 (8.28%)") + ggtitle("Nuclear DNA") 

p1
```

### Mitochondrial - Figure 1D (right hand side)

```{r, fig.height=5, fig.width=10}

rc = t(as.matrix(mito[,c(5:19)]/100))
rc = rc[,colSums(rc > 0) >= 3] # choose only those SNPs where minor allele freq is not equal to 0 in at least 3 pools
pc = prcomp(rc, center = TRUE, scale. = TRUE)
summ <- summary(pc)
summ
rownames(pc$x) = c("Alexandrina1", "Alexandrina2", "Alexandrina3", "Mapourika1", "Mapourika2", "Mapourika3", "Middleton1", "Middleton2", "Middleton3", "Paringa1", "Paringa2", "Paringa3", "Selfe1", "Selfe2", "Selfe3")
rownames(metadata) = c("Alexandrina1", "Alexandrina2", "Alexandrina3", "Mapourika1", "Mapourika2", "Mapourika3", "Middleton1", "Middleton2", "Middleton3", "Paringa1", "Paringa2", "Paringa3", "Paringa4", "Selfe1", "Selfe2", "Selfe3")
df = pc$x %>% data.frame() %>% rownames_to_column("pool") %>% left_join(metadata[-13, , drop = FALSE] %>% rownames_to_column("pool"), by = "pool") 
p2 = df %>% ggplot(aes(PC1, PC2, color = lake)) + geom_point(size = 3) + theme_bw() + scale_color_manual(values = colors$lake) + geom_label_repel(aes(label = pool)) + labs(y = "PC2 (17.5%)",x = "PC1 (45%)" ) + ggtitle("Mitochondrial DNA") + theme(legend.position = "")  

pdf(paste0(working_dir,"/../Figures/PCA_NandMT.v2.pdf"), height = 4, width = 10)
p1 + p2 + plot_layout(guides = "collect")
dev.off()
```

## Covariance matrix from BayPass - Supplementary Figure 1

```{r, fig.height=5, fig.width=10}

pdf(paste0(working_dir,"/../Figures/Baypass_matrix.pdf"), height = 7, width = 10)
labels = c(metadata[-13,])
rownames(omega_cov_matrix) = c("Alexandrina1", "Alexandrina2", "Alexandrina3", "Mapourika1", "Mapourika2", "Mapourika3", "Middleton1", "Middleton2", "Middleton3", "Paringa1", "Paringa2", "Paringa3", "Selfe1", "Selfe2", "Selfe3")
colnames(omega_cov_matrix) = c("Alexandrina1", "Alexandrina2", "Alexandrina3", "Mapourika1", "Mapourika2", "Mapourika3", "Middleton1", "Middleton2", "Middleton3", "Paringa1", "Paringa2", "Paringa3", "Selfe1", "Selfe2", "Selfe3")
ht = Heatmap(omega_cov_matrix, 
        heatmap_legend_param = list(direction = "horizontal"),
        row_names_gp = gpar(fontsize = 10),
        row_split  = labels, 
        column_split = labels,  
        border = T, 
        top_annotation = HeatmapAnnotation(empty = anno_empty(border = FALSE),
                                           foo = anno_block(gp = gpar(fill = c("darkgreen", "gold", "orange", "purple", "cornflowerblue")),
                                                            labels = c("Selfe", "Middleton", "Alexandrina", "Paringa","Mapourika"))), 
        show_row_dend = F, 
        show_column_names = F, 
        circlize::colorRamp2(c(min(omega_cov_matrix), 0.9, max(omega_cov_matrix)), c("blue", "white", "red")),
        name = "Similarity")

draw(ht ,heatmap_legend_side = "bottom")
dev.off()
```

Additional mitochondrial not in manuscript

```{r, fig.height=5, fig.width=10}
labels = c(metadata[-13,])
ht = Heatmap(omega_cov_matrix_mt, 
             heatmap_legend_param = list(direction = "horizontal"),
             row_names_gp = gpar(fontsize = 10),
             row_split  = labels, 
             column_split = labels,  
             border = T, 
             top_annotation = HeatmapAnnotation(empty = anno_empty(border = FALSE),
                                                foo = anno_block(gp = gpar(fill = c("darkgreen", "purple", "cornflowerblue", "orange", "gold")),
                                                                 labels = c("Selfe", "Paringa","Mapourika", "Alexandrina","Middleton"))), 
             show_row_dend = F, 
             show_column_names = F, 
             circlize::colorRamp2(c(min(omega_cov_matrix_mt), 0.9, max(omega_cov_matrix_mt)), c("blue", "white", "red")),
             name = "Similarity")

draw(ht ,heatmap_legend_side = "bottom")
```
